---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<img src="man/figures/logo.svg" align="right" height="180" alt="redcapmissing hex logo" />

# redcapmissing

<!-- badges: start -->
![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-339999)
![License](https://img.shields.io/badge/license-MIT-blue.svg)
<!-- badges: end -->

`redcapmissing` builds branching-aware missingness reports for REDCap record exports.

It is designed for REDCap data-quality workflows where expected values depend on project structure, including:

- branching logic
- longitudinal event mapping
- repeating instruments
- checkbox semantics
- whole-form missingness
- `pointblank` validation summaries and failed-row extracts

## Dependency on `redcapAPI`

`redcapmissing` is built on top of [`redcapAPI`](https://github.com/vubiostat/redcapAPI).
In real workflows, `rcon` is not an arbitrary list-like object: it is ordinarily a
`redcapAPI::redcapConnection()` object, and `data` is ordinarily created with
`redcapAPI::exportRecordsTyped()`.

This package depends on `redcapAPI` for REDCap metadata access, form-event
mapping, repeating-instrument structure, and REDCap-style blank-value handling.

## Why use `redcapmissing`?

`redcapmissing` separates REDCap-aware expectation building from auditable validation output:

- `redcapAPI` exposes the REDCap project metadata and structural context.
- R code determines which record, event, repeat, and field contexts should be checked.
- `pointblank` records the validation plan and exposes scope-specific summaries.
- Row-level failures remain easy to extract for correction workflows.

## Installation

```r
# install.packages("pak")
pak::pkg_install("blankuzr/redcapmissing")
```

Installing `redcapmissing` also installs `redcapAPI` as a package dependency.

## Core functions

- `redcap_missing_report()`
  - build a missingness report for one REDCap form/instrument
- `redcap_missing_summary()`
  - format the `pointblank` validation summary from a report object

## What the report returns

`redcap_missing_report()` returns a structured list, not just a single table. The
most commonly used components are:

- `report$missing`
  - the row-level failed extract, intended as the main correction queue
- `report$agent`
  - the interrogated `pointblank` agent, including validation metadata and the
    underlying summary counts
- `report$expected`
  - the expected field-level rows that were actually assessed after branching
    logic and filtering
- `report$form_missing`, `report$event_missing`, `report$repeat_missing`
  - the failed rows for the three whole-context missingness scopes
- `report$form_checks`, `report$event_checks`, `report$repeat_checks`
  - the full scope-specific check tables, including both passing and failing
    contexts

This separation matters because REDCap exports can fail at different levels:
missing event rows, missing repeat-instance rows, wholly blank form rows, and
ordinary field-level missingness are not the same problem.

## Summary helper

`redcap_missing_summary()` is a convenience formatter for the `pointblank`
summary stored inside `report$agent`. It does not replace the row-level extracts;
it gives a clean evaluation table for reporting and review.

It returns:

- `agent_summary`
  - a `flextable` object
- `agent_summary_html`
  - an HTML representation of the same summary table

```{r, eval = FALSE}
summary_tbl <- redcap_missing_summary(report)
summary_tbl$agent_summary
summary_tbl$agent_summary_html
```

## Example

The example below uses a lightweight synthetic stand-in for a REDCap connection
so it can run without live REDCap access. In production use, create `rcon` with
`redcapAPI::redcapConnection()` and export records with
`redcapAPI::exportRecordsTyped()`.

```{r, eval = FALSE}
library(redcapmissing)

metadata <- tibble::tibble(
  field_name = c("record_id", "branch_flag", "required_note", "conditional_note"),
  form_name = c("baseline_form", "baseline_form", "baseline_form", "baseline_form"),
  field_type = c("text", "yesno", "text", "text"),
  field_label = c("Record ID", "Branch flag", "Required note", "Conditional note"),
  select_choices_or_calculations = c("", "", "", ""),
  text_validation_type_or_show_slider_number = c("", "", "", ""),
  branching_logic = c("", "", "", "[branch_flag] = '1'"),
  required_field = c("y", "y", "y", "y")
)

rcon <- list(metadata = function() metadata)

records <- tibble::tibble(
  record_id = c("r1", "r2"),
  branch_flag = c("1", "0"),
  required_note = c("entered", "entered"),
  conditional_note = c("", "")
)

report <- redcap_missing_report(
  data = records,
  rcon = rcon,
  form = "baseline_form"
)

report$missing[, c("record_id", "field_name", "missing_scope")]
```

## Missingness scopes

REDCap exports make it important to distinguish between a missing **value** and a
missing **row context**. In longitudinal and repeating projects, a record can be
missing because REDCap exported no row at all for the relevant event or repeat
instance. In other contexts, REDCap does export a row, but every field on the
form is still blank. `redcapmissing` separates those cases into four scopes.

### `event_absent`

Use this scope when a form is offered on a longitudinal event, but the export has
no row at all for that record-event context.

Why this exists:

- REDCap exports cannot show field-level blanks for an event row that was never exported.
- The package therefore uses the project form-event mapping from `rcon` to build
  expected record-event contexts before any field-level check can happen.
- When an expected event row is absent, the report returns one row for the
  missing event context instead of many synthetic field failures.

### `repeat_absent`

Use this scope when a form is a repeating instrument and an expected repeat
instance row does not exist in the export.

Why this exists:

- REDCap only exports repeat instances that actually exist.
- If instance 2 should exist but was never created, there is no exported row to inspect.
- `expected_repeats` therefore acts as an expected-row rule: the function builds
  the expected record-event-repeat contexts and compares them to what REDCap exported.
- When an expected repeat row is absent, the report returns one row for that
  missing repeat context.

### `form_blank`

Use this scope when REDCap exported the row for the form context, but every
data-capturing field on that form is blank or unchecked.

Why this exists:

- In practice, this often means the form was available in REDCap but no real data entry started.
- REDCap may still export the record/event/repeat row even though every form
  field is empty.
- Reporting each field separately would overstate the problem, so the package
  records a single form-level failure for that context.

### `field`

Use this scope when the row context exists, the form is not wholly blank, and a
specific field is expected after REDCap branching logic is evaluated.

Why this exists:

- This is the ordinary field-level missingness check.
- A field is only assessed after the package confirms the row context exists,
  the form is not wholly blank, the field is on the requested form, and its
  branching logic is open.
- This is where REDCap-specific missingness behaves most like standard value-level QA.

## Restricting assessment to selected events

When a form is offered on many REDCap events, you can restrict assessment to a
chosen subset with `desired_events`. If you do not supply it, the function
defaults to all REDCap events where the form is offered.

```r
followup_report <- redcap_missing_report(
  data = records,
  rcon = rcon,
  form = "patient_status",
  desired_events = c(
    "follow_up_1_arm_1",
    "follow_up_2_arm_1",
    "follow_up_3_arm_1"
  )
)
```

This is especially useful in longitudinal REDCap projects where several events
play the same conceptual role but only a subset should count toward the current
missingness review.

If a form is regular on some requested events and repeating on others, the
package applies scopes by event type:

- regular-form events use the standard `field`, `form_blank`, and `event_absent` logic
- repeating-instrument events use `field`, `form_blank`, and `repeat_absent` logic

When `expected_repeats` is omitted, the default `1L` assumption is only applied
for the requested events where the form actually repeats.

## Repeat expectations

For repeating instruments, `expected_repeats` applies a uniform expectation to all assessed record/event contexts.

```r
repeat_report <- redcap_missing_report(
  data = records,
  rcon = rcon,
  form = "repeat_form",
  expected_repeats = 2L
)
```

This checks that repeat instances `1` and `2` exist everywhere that `repeat_form` is expected.
The key REDCap detail is that missing repeat instances are absent as rows, not merely
blank as values. `expected_repeats` lets `redcapmissing` create those expected
row contexts explicitly before comparing them to the export.

## Acknowledgement and citation

This package relies heavily on `redcapAPI` and would not be practical without it.
If `redcapmissing` contributes to your work, please also cite `redcapAPI`.

### Foundational package citation

> Nutter B, Garbett S, Obregon S, Obadia T, Lehr M, High B, Lane S, Beasley W,
> Gray W, Kennedy N, Hsi-Nien T, Horner J, Stephens J, Beck C, Johnson B,
> Chase P, Tobias P (2026). *redcapAPI: Accessing data from REDCap projects
> using the API*. R package version 2.12.0.
> <https://doi.org/10.5281/zenodo.10564837>

### Current stewardship and project resources

Current public stewardship appears under VUMC Biostatistics / `vubiostat`, with
Shawn Garbett listed as maintainer in current package documentation and the
upstream project README stating that ownership transfer to VUMC Biostatistics is
complete.

Useful current references:

- VUMC Biostatistics redcapAPI project page and abstract by Savannah Obregon,
  Shawn Garbett, and Benjamin Nutter:
  <https://www.vumc.org/biostatistics/node/565>
- Current GitHub repository for the package:
  <https://github.com/vubiostat/redcapAPI>
- Current package site:
  <https://vubiostat.r-universe.dev/redcapAPI>

## Learn more

See the package vignette for a fuller synthetic walk-through of branching-aware and repeat-aware validation.

## Development

```r
devtools::document()
devtools::test()
devtools::check()
```
