---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
options(width = 120)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# redcapmissing

<!-- badges: start -->
![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-339999)
![License](https://img.shields.io/badge/license-MIT-blue.svg)
<!-- badges: end -->

`redcapmissing` builds branching-aware missingness reports for REDCap record
exports. It combines `redcapAPI` project metadata with typed REDCap exports so
missing values, absent event rows, absent repeat instances, blank forms, and
field-specific gaps stay separate in the report.

<p align="center">
  <img src="man/figures/logo.svg" width="160" alt="redcapmissing hex logo" />
</p>

## Installation

`redcapmissing` requires R 4.1.0 or later. Install the current GitHub version
with:

```r
# install.packages("pak")
pak::pak("blankuzr/redcapmissing")
```

## REDCap inputs

In production workflows, create a `redcapAPI::redcapConnection()` object and
export records with `redcapAPI::exportRecordsTyped()`. Use coded values for
branching-logic fields and raw values for checkbox and system fields:

```r
records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    truefalse = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)
```

## Validation model

`redcapmissing` uses six validation checks across two conceptual
validation-levels. Failed `on-route` checks remove that record/event/repeat/form
context from every downstream check. The event/form validation-level is written
`event:form / event:form:instance` because the emitted report row resolves to
`event:form` when the requested form is assessed in a non-repeating event
context and `event:form:instance` when it is assessed in a repeat-instance
context. The `event` level is a summary layer. `form-complete` and
`event-complete` are `detour` checks: they report failures without feeding back
into the pass-only pipeline.

```mermaid
flowchart TB
  A["event-row-started<br/>non-repeating context<br/>on-route"] ==>|"PASS only"| C["form-started<br/>event:form / event:form:instance<br/>on-route"]
  B["instance-row-started<br/>repeat-instance context<br/>on-route"] ==>|"PASS only"| C
  C ==>|"PASS only"| E["field-complete<br/>event:form / event:form:instance<br/>on-route"]
  C -. "detour: reports only" .-> D["form-complete<br/>event:form / event:form:instance<br/>detour"]

  A -. "on-route result" .-> O(("on-route<br/>results by event"))
  B -. "on-route result" .-> O
  C -. "on-route result" .-> O
  E -. "on-route result" .-> O
  O -. "detour rollup" .-> F["event-complete<br/>event / detour"]

  classDef onRoute fill:#eaf5ee,stroke:#258457,color:#173b2b,stroke-width:2px;
  classDef detour fill:#fff4d7,stroke:#c77900,color:#513300,stroke-width:2px;
  classDef rollup fill:#e7f0ff,stroke:#3569a8,color:#183353,stroke-width:2px;
  class A,B,C,E onRoute;
  class D,F detour;
  class O rollup;
```

| validation_level | validation_check | validation_check_type | Meaning |
|---|---|---|---|
| event:form / event:form:instance | `event-row-started` | `on-route` | The expected REDCap event row exists in the export. |
| event:form / event:form:instance | `instance-row-started` | `on-route` | The expected REDCap repeat instance row exists in the export. |
| event:form / event:form:instance | `form-started` | `on-route` | The exported form context has at least one entered data-capturing field. |
| event:form / event:form:instance | `form-complete` | `detour` | All expected fields are complete for an evaluable form context. |
| event:form / event:form:instance | `field-complete` | `on-route` | A specific expected field is complete after branching and filtering. |
| event | `event-complete` | `detour` | All on-route checks pass within the REDCap event context. |

Call `redcapmissing::registry()` to inspect the package registry that drives
this model:

```{r, eval = FALSE}
redcapmissing::registry()
```

## Minimal workflow

The example below uses a synthetic connection-like object so it can run without
live REDCap credentials. In routine use, replace `rcon` and `records` with a
real `redcapAPI::redcapConnection()` and typed export.

```{r, message = FALSE}
library(redcapmissing)

metadata <- tibble::tibble(
  field_name = c("record_id", "branch_flag", "required_note", "conditional_note"),
  form_name = "baseline_form",
  field_type = c("text", "yesno", "text", "text"),
  field_label = c("Record ID", "Branch flag", "Required note", "Conditional note"),
  select_choices_or_calculations = "",
  text_validation_type_or_show_slider_number = "",
  branching_logic = c("", "", "", "[branch_flag] = '1'"),
  required_field = c("y", "y", "y", "y")
)

rcon <- list(
  metadata = function() metadata,
  instruments = function() tibble::tibble(
    instrument_name = "baseline_form",
    instrument_label = "Baseline form"
  )
)

records <- tibble::tibble(
  record_id = c("r1", "r2"),
  branch_flag = c("1", "0"),
  required_note = c("entered", "entered"),
  conditional_note = c("", "")
)

report <- find_missing(
  data = records,
  rcon = rcon,
  forms = "baseline_form"
)

missing_rows <- as.data.frame(report$missing[
  ,
  c(
    "record_id",
    "validation_level",
    "validation_check_type",
    "validation_check",
    "field_name"
  )
])
print(missing_rows, row.names = FALSE)

summary <- tidy(report)
summary_rows <- as.data.frame(summary[
  summary$assessed > 0,
  c(
    "validation_level",
    "validation_check",
    "validation_check_type",
    "assessed",
    "failed"
  )
])
print(summary_rows, row.names = FALSE)
```

## Report outputs

`find_missing()` returns a `"redcapmissing"` object centered on an interrogated
`pointblank` agent. Common outputs are:

- `report$missing`: failed rows extracted from `pointblank`
- `report$validation_rows`: all rows supplied to `pointblank`
- `report$event_row_started_failures`, `report$instance_row_started_failures`,
  `report$form_started_failures`, `report$form_complete_failures`,
  `report$field_complete_failures`, and `report$event_complete_failures`:
  check-specific failure tables
- `tidy(report)`: one summary row per validation check and REDCap context

`tidy(report)` returns raw REDCap event/form context first, followed by
canonical validation columns and pass/fail rates. Repeat instrument and repeat
instance columns are included only when the report contains repeat contexts.
`validation_level` is `event:form`, `event:form:instance`, or `event`, and
`validation_check_type` is `on-route` or `detour`.

Optional reporting helpers are available for formatted outputs.
`flex(report)` requires `flextable` and `glue`; `flex_html(flex(report))`
also requires `htmltools`. `flex(report)` displays labeled event, form, repeat
context, validation-check, and pass/fail columns. Use raw REDCap values in
`flex(events = ...)` and `flex(forms = ...)` to subset rows before display
labels are applied.

## Events and repeats

Use `events` to restrict multi-event forms to selected REDCap events. Use
`instances` to declare expected repeat instances when REDCap would otherwise
omit nonexistent repeat rows from the export.

For a repeating form, declare the expected repeat instances after exporting
records and creating the REDCap connection object:

```{r, eval = FALSE}
repeat_report <- find_missing(
  data = typed_records,
  rcon = rcon,
  forms = "repeat_form",
  instances = 2L
)
```

For forms that are regular on some requested events and repeating on others,
the package activates event/form checks per context: regular event contexts use
`event-row-started`, repeating contexts use `instance-row-started`, and repeat
instances roll up to their parent event for `event-complete`.

See the [package vignette](vignettes/redcapmissing.Rmd#repeat-expectations) for
a runnable synthetic repeat example.

## Acknowledgement and citation

### `redcapAPI`

This package relies heavily on `redcapAPI` and would not be practical without
it. If `redcapmissing` contributes to your work, please also cite
`redcapAPI`.

#### Foundational package citation

> Nutter B, Garbett S, Obregon S, Obadia T, Lehr M, High B, Lane S,
> Beasley W, Gray W, Kennedy N, Hsi-Nien T, Horner J, Stephens J, Beck C,
> Johnson B, Chase P, Tobias P (2026). *redcapAPI: Accessing data from REDCap
> projects using the API*. R package version 2.12.0.
> <https://doi.org/10.5281/zenodo.10564837>

#### Current package ownership and maintenance

Current public stewardship appears under VUMC Biostatistics / `vubiostat`, with
Shawn Garbett listed as maintainer in current package documentation and the
upstream project README stating that ownership transfer to VUMC Biostatistics is
complete.

Useful current `redcapAPI` references:

- VUMC Biostatistics redcapAPI project page and abstract by Savannah Obregon,
  Shawn Garbett, and Benjamin Nutter:
  <https://www.vumc.org/biostatistics/node/565>
- Current GitHub repository for the package:
  <https://github.com/vubiostat/redcapAPI>
- Current package site:
  <https://vubiostat.r-universe.dev/redcapAPI>

### `pointblank`

This package relies on `pointblank` for validating missingness, standardizing
return summaries, and exposing failed-row details.

#### Current package resources

Useful current `pointblank` references:

- R package GitHub repository:
  <https://github.com/rstudio/pointblank>
- R package site:
  <https://rstudio.github.io/pointblank/>
- Python package GitHub repository:
  <https://github.com/posit-dev/pointblank>
- Python package site:
  <https://posit-dev.github.io/pointblank/>

## Learn more

- Run `vignette("redcapmissing", package = "redcapmissing")` after installing
  the package.
- Read the [getting started vignette](vignettes/redcapmissing.Rmd) for a fuller
  synthetic walk-through of branching-aware, repeat-aware, and
  validation-registry-driven missingness reporting.
- Review [NEWS.md](NEWS.md) for user-facing changes by version.
- Report bugs and feature requests in
  [GitHub issues](https://github.com/blankuzr/redcapmissing/issues).

## Development

```r
devtools::document()
devtools::test()
devtools::check()
```
