---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
options(width = 120)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  default_output_hook(gsub("[[:blank:]]+(?=\\r?\\n|$)", "", x, perl = TRUE), options)
})
```

# redcapmissing

<!-- badges: start -->
![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-339999)
![License](https://img.shields.io/badge/license-MIT-blue.svg)
<!-- badges: end -->

`redcapmissing` builds branching-aware missingness reports from REDCap record
exports. It distinguishes missing field values from absent event rows, absent
repeat instances, and forms that were never started, while respecting REDCap
branching logic and project structure.

<p align="center">
  <img src="man/figures/logo.svg" width="160" alt="redcapmissing hex logo" />
</p>

## Installation

`redcapmissing` requires R 4.1.0 or later. Install it from GitHub with:

```r
# install.packages("pak")
pak::pak("blankuzr/redcapmissing")
```

## REDCap inputs

For a live project, create a `redcapAPI::redcapConnection()` and export records
with `redcapAPI::exportRecordsTyped()`. Branching-logic fields must remain in
REDCap code space, while checkbox and system fields should remain raw. This
keeps values aligned with the codes used in REDCap logic and preserves the
system columns needed to identify events and repeat instances.

```r
records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    truefalse = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)
```

## First report

This synthetic example runs without REDCap credentials. The connection-like
object supplies the same metadata and instrument methods used from a live
connection, and the record values use REDCap codes.

```{r first-report, message = FALSE}
library(redcapmissing)

metadata <- tibble::tibble(
  field_name = c("record_id", "branch_flag", "required_note", "conditional_note"),
  form_name = "baseline_form",
  field_type = c("text", "yesno", "text", "text"),
  field_label = c("Record ID", "Branch flag", "Required note", "Conditional note"),
  select_choices_or_calculations = "",
  text_validation_type_or_show_slider_number = "",
  branching_logic = c("", "", "", "[branch_flag] = '1'"),
  required_field = "y"
)

rcon <- list(
  metadata = function() metadata,
  instruments = function() tibble::tibble(
    instrument_name = "baseline_form",
    instrument_label = "Baseline form"
  )
)

records <- tibble::tibble(
  record_id = c("r1", "r2"),
  branch_flag = c("1", "0"),
  required_note = "entered",
  conditional_note = ""
)

report <- find_missing(
  data = records,
  rcon = rcon,
  forms = "baseline_form"
)

get_summary(report)[
  ,
  c("validation_check", "assessed", "failed")
]

get_missing(
  report,
  validation_check = "field-complete"
)[, c("record_id", "form", "field_name")]
```

Record `r1` has an open branch for `conditional_note`, so its blank value is a
`field-complete` failure. The same branch is closed for `r2`, so that field is
not assessed for the second record.

## How validation works

Checks run in registry order for each record, event, repeat, and form context:

| Check | What it establishes |
|---|---|
| `event-row-started` | The expected REDCap event row exists in the export. |
| `instance-row-started` | The expected repeat-instance row exists in the export. |
| `form-started` | The form context contains at least one entered data-capturing field. |
| `field-complete` | Each applicable field is answered after branching logic and field selection are applied. |

Blank strings and `NA` values are missing only when the field is expected in
that context. A checkbox question is answered when at least one exported child
choice is selected; its unselected siblings are not separate missing fields.
By default, `find_missing()` assesses required fields and excludes descriptive
fields. Use `required_fields`, `exclude_types`, and `ignore_fields` when the
report needs a different field scope.

When a check fails, that context is removed from every downstream check. An
absent event row, for example, is reported at `event-row-started` instead of
being counted again as a form that was not started and as several missing
fields. Use `registry()` to inspect the validation checks and their labels.

## Explore and format results

Use `get_summary()` for validation counts and rates and `get_missing()` for the
failed record or field rows. Both accessors can filter by validation check,
event, or form; their reference topics document the complete schemas and
filtering rules.

```{r format-results, eval = FALSE}
field_summary <- get_summary(
  report,
  validation_check = "field-complete"
)

field_missing <- get_missing(
  report,
  validation_check = "field-complete"
)

flexify(field_summary)
flexify(field_missing)
flex_event_forms(report)
flex_html(flexify(field_summary))
```

`flexify()` formats an accessor result as a flextable,
`flex_event_forms()` creates a reduced event/form view, and `flex_html()`
serializes either kind of flextable for HTML workflows. Filter and select rows
or columns before formatting when a smaller table is needed.

## Events, records, and repeats

The main scoping arguments to `find_missing()` answer different questions:

- `events` selects the REDCap events on which a requested form should be
  assessed, including events that REDCap may omit entirely from an export
  because no row was started.
- `records` restricts record eligibility by raw event name and, when needed,
  by form or repeat instance.
- `instances` declares expected repeat-instance IDs, including instances that
  REDCap may omit entirely from an export because no row was started.

Forms that are regular on some events and repeating on others are assessed in
their actual context: regular event rows use `event-row-started`, while repeat
rows use `instance-row-started`. See `?find_missing` for the accepted argument
shapes and the package vignette for runnable longitudinal and repeat examples.

## Learn more

- Read the [getting started vignette](vignettes/redcapmissing.Rmd) for branching,
  event, record-eligibility, and repeat-instance examples; after installation,
  open the local copy with `vignette("redcapmissing", package = "redcapmissing")`.
- Open `?find_missing`, `?get_summary`, `?get_missing`, `?flexify`, or
  `?flex_event_forms` for the full function contracts.
- Review [NEWS.md](NEWS.md) for changes by version.
- Report bugs and request features in
  [GitHub issues](https://github.com/blankuzr/redcapmissing/issues).
