---
title: "Getting started with redcapmissing"
vignette: >
  %\VignetteIndexEntry{Getting started with redcapmissing}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(redcapmissing)
```

`redcapmissing` builds branching-aware missingness reports for REDCap record
exports. In real workflows, callers create a
`redcapAPI::redcapConnection()`, export records with
`redcapAPI::exportRecordsTyped()`, and pass both objects into `find_missing()`.

The examples below are synthetic so the vignette can run without live REDCap
credentials.

## Validation registry

The package validation model is available through `registry()`:

```{r}
registry()
```

The checks are organized by validation level and validation-check type:

- `event:form / event:form:instance`: one contextual event/form
  validation-level. Report rows use `event:form` for non-repeating contexts and
  `event:form:instance` for repeat-instance contexts.
- `event`: event-level rollups.

`event-row-started`, `instance-row-started`, `form-started`, and
`field-complete` are `on-route`: failures remove that context from downstream
assessment. `form-complete` and `event-complete` are `detour`: failures are
reported without blocking later checks. `event-complete` has the display
meaning `all forms on event complete` and is computed from event-level
downstream-gating `on-route` checks; it does not count `form-complete` rows
because `form-complete` is also a `detour`.

| validation_level | validation_check | validation_check_type | Meaning |
|---|---|---|---|
| event:form / event:form:instance | `event-row-started` | `on-route` | The expected REDCap event row exists in the export. |
| event:form / event:form:instance | `instance-row-started` | `on-route` | The expected REDCap repeat instance row exists in the export. |
| event:form / event:form:instance | `form-started` | `on-route` | The exported form context has at least one entered data-capturing field. |
| event:form / event:form:instance | `form-complete` | `detour` | all form fields complete |
| event:form / event:form:instance | `field-complete` | `on-route` | field complete |
| event | `event-complete` | `detour` | all forms on event complete |

## Synthetic REDCap helpers

```{r}
meta_row <- function(
  field_name,
  form_name,
  field_type = "text",
  field_label = field_name,
  choices = "",
  branching = "",
  required = ""
) {
  tibble::tibble(
    field_name = field_name,
    form_name = form_name,
    field_type = field_type,
    field_label = field_label,
    select_choices_or_calculations = choices,
    text_validation_type_or_show_slider_number = "",
    branching_logic = branching,
    required_field = required
  )
}

fake_rcon <- function(metadata, mapping = NULL, repeat_instrument_event = NULL) {
  forms <- unique(as.character(metadata$form_name))
  instruments <- tibble::tibble(
    instrument_name = forms,
    instrument_label = paste(forms, "label")
  )

  list(
    metadata = function() metadata,
    instruments = function() instruments,
    mapping = function() mapping,
    mappings = function() mapping,
    repeatInstrumentEvent = function() repeat_instrument_event
  )
}
```

## Real `redcapAPI` workflow

For real REDCap projects, keep branching-logic fields in REDCap code space and
leave checkbox and system fields raw:

```{r, eval = FALSE}
rcon <- redcapAPI::redcapConnection(
  url = Sys.getenv("REDCAP_API_URL"),
  token = Sys.getenv("REDCAP_API_TOKEN")
)

records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    truefalse = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)

report <- find_missing(
  data = records,
  rcon = rcon,
  forms = "baseline_form"
)
```

## Branching-aware field checks

```{r}
metadata <- dplyr::bind_rows(
  meta_row("record_id", "baseline_form", required = "y"),
  meta_row("branch_flag", "baseline_form", field_type = "yesno", required = "y"),
  meta_row("required_note", "baseline_form", required = "y"),
  meta_row(
    "conditional_note",
    "baseline_form",
    branching = "[branch_flag] = '1'",
    required = "y"
  )
)

records <- tibble::tibble(
  record_id = c("r1", "r2"),
  branch_flag = c("1", "0"),
  required_note = c("entered", "entered"),
  conditional_note = c("", "")
)

report <- find_missing(
  data = records,
  rcon = fake_rcon(metadata),
  forms = "baseline_form"
)

report$missing[, c(
  "record_id",
  "validation_level",
  "validation_check_type",
  "validation_check",
  "field_name"
)]
```

Record `r1` is branch-open for `conditional_note`, so the missing value is a
`field-complete` failure. Record `r2` is branch-closed, so the field is not in
the assessment pool.

## Event-row gating

REDCap exports may omit entire event rows. `event-row-started` catches that
before form and field checks run downstream.

```{r}
status_metadata <- dplyr::bind_rows(
  meta_row("record_id", "status_form", required = "y"),
  meta_row("status_started", "status_form", required = "y"),
  meta_row("status_value", "status_form", required = "y")
)

mapping <- tibble::tibble(
  arm_num = c(1, 1, 1),
  unique_event_name = c("event_1_arm_1", "event_2_arm_1", "event_3_arm_1"),
  form = c("status_form", "status_form", "status_form")
)

event_records <- tibble::tibble(
  record_id = c("r1", "r1", "r2", "r2"),
  redcap_event_name = c(
    "event_1_arm_1",
    "event_2_arm_1",
    "event_1_arm_1",
    "event_2_arm_1"
  ),
  status_started = c("yes", "yes", "yes", "yes"),
  status_value = c("entered", "", "entered", "entered")
)

event_report <- find_missing(
  data = event_records,
  rcon = fake_rcon(status_metadata, mapping = mapping),
  forms = "status_form"
)

tidy(event_report)[
  ,
  c("validation_level", "validation_check", "redcap_event_name", "assessed", "failed")
]
```

`event_3_arm_1` fails `event-row-started`, so it is not assessed by
`form-started`, `form-complete`, or `field-complete`. Because `event-complete`
rolls up on-route failures, the same record/event failures are also visible in
the event-level summary.

## Event-specific record eligibility

Use `records` when records become eligible for validation on different REDCap
events at different times. The argument is a named list keyed by raw
`redcap_event_name`. Each non-empty entry limits that event to the listed
record IDs. Events not named in `records` are still checked when the requested
form is offered on that event: every applicable non-ignored record in `data` is
considered, and records with no exported row for that event fail
`event-row-started` before downstream form or field checks. Empty entries behave
like omitted events. To remove an event from assessment entirely, exclude it
with `events` rather than omitting it from `records`.

```{r}
staged_metadata <- dplyr::bind_rows(
  meta_row("record_id", "demographics", required = "y"),
  meta_row("demographics_started", "demographics", required = "y"),
  meta_row("demographics_value", "demographics", required = "y"),
  meta_row("surgery_started", "surgery", required = "y"),
  meta_row("surgery_value", "surgery", required = "y")
)

staged_mapping <- tibble::tibble(
  arm_num = c(1, 1, 1),
  unique_event_name = c("event_a", "event_b", "event_c"),
  form = c("demographics", "surgery", "surgery")
)

staged_records <- tibble::tibble(
  record_id = c("record_a", "record_a", "record_b", "record_b", "record_b"),
  redcap_event_name = c("event_a", "event_b", "event_a", "event_b", "event_c"),
  demographics_started = c("yes", "", "yes", "", ""),
  demographics_value = c("entered", "", "entered", "", ""),
  surgery_started = c("", "yes", "", "yes", "yes"),
  surgery_value = c("", "entered", "", "entered", "")
)

staged_report <- find_missing(
  data = staged_records,
  rcon = fake_rcon(staged_metadata, mapping = staged_mapping),
  forms = c("surgery", "demographics"),
  records = list(
    event_a = c("record_a", "record_b"),
    event_b = c("record_a", "record_b"),
    event_c = "record_b"
  )
)

staged_report$spec$eligible_records

staged_summary <- tidy(staged_report)
staged_summary[
  staged_summary$validation_check == "event-row-started",
  c("form", "redcap_event_name", "assessed", "failed")
]
```

Record `record_a` is assessed for `surgery` on `event_b`, but not on
`event_c`. Record `record_b` is assessed on `event_c`, where its blank surgery
value fails downstream field completeness.

If another requested event offered `demographics` and was not named in
`records`, the `records` argument would not limit that event. The package would
consider every applicable non-ignored record in `data` for that event, then use
`event-row-started` to decide whether each record actually has an event row.

## Repeat expectations

REDCap does not export nonexistent repeat rows. Use `instances` to declare the
repeat instances expected for the report.

```{r}
repeat_metadata <- dplyr::bind_rows(
  meta_row("record_id", "screen_form", required = "y"),
  meta_row("screen_started", "screen_form", required = "y"),
  meta_row("repeat_started", "repeat_form", required = "y"),
  meta_row("repeat_value", "repeat_form", required = "y")
)

repeat_mapping <- tibble::tibble(
  arm_num = c(1, 1),
  unique_event_name = c("baseline_event", "baseline_event"),
  form = c("screen_form", "repeat_form")
)

repeat_map <- tibble::tibble(
  event_name = "baseline_event",
  form_name = "repeat_form",
  custom_form_label = ""
)

repeat_records <- tibble::tibble(
  record_id = c("r1", "r2", "r2"),
  redcap_event_name = c("baseline_event", "baseline_event", "baseline_event"),
  redcap_repeat_instrument = c("repeat_form", "repeat_form", "repeat_form"),
  redcap_repeat_instance = c("1", "1", "2"),
  repeat_started = c("yes", "yes", "yes"),
  repeat_value = c("10", "20", "30")
)

repeat_report <- find_missing(
  data = repeat_records,
  rcon = fake_rcon(
    repeat_metadata,
    mapping = repeat_mapping,
    repeat_instrument_event = repeat_map
  ),
  forms = "repeat_form",
  instances = 2L
)

repeat_report$missing[
  repeat_report$missing$validation_check == "instance-row-started",
  c("record_id", "redcap_repeat_instance", "validation_check")
]
```

The missing instance `2` for record `r1` fails `instance-row-started` and is
removed from downstream form and field assessment. It also causes record `r1`
to fail `event-complete` for `baseline_event` because repeat-instance on-route
failures belong to the parent REDCap event.

## Mixed regular and repeating contexts

Some REDCap projects offer the same form as regular on one event and repeating
on another. The package activates event/form checks per context:

- regular event contexts use `event-row-started`
- repeating contexts use `instance-row-started`
- both context types roll up into `event-complete`

If `events` excludes the repeating event, `instances` is ignored for that form.
If a requested repeating context omits `instances`, the package assumes
instance `1` and warns once.

## Inspecting results

`tidy(report)` returns a raw context-first summary with canonical validation
metadata. Repeat context columns are shown only when the report contains repeat
contexts:

```{r}
tidy(report)
```

For reporting workflows, optional formatting helpers are available:

```{r, eval = FALSE}
ft <- flex(report)
field_ft <- flex(report, validation_check = "field-complete")
event_form_ft <- flex_event_forms(report)
html <- flex_html(ft)
```

`flex()` displays the canonical checks with human-readable event, form, and
repeat labels. Its `events`, `forms`, and `validation_check` arguments use raw
values from `tidy(report)` to filter rows before labels are applied. `tidy()`
and `report$summary` keep the canonical hyphenated `validation_check` values.
`flex_event_forms()` returns a reduced event/form table with total record N in
the N column label, event row-started passed/assessed counts on event rows,
form-complete counts, and fields-missing counts nested under each event. Repeat
form rows show repeat-instance passed/assessed counts in the N column, while
non-repeat form rows leave that cell blank. Projects whose REDCap record ID
field has another name still produce populated event and form rows. The public
`validation_level` values are `event:form`, `event:form:instance`, and `event`.

## Citation and acknowledgement

`redcapmissing` depends heavily on `redcapAPI`. Please cite `redcapAPI` when
`redcapmissing` is part of a REDCap workflow.

### Foundational package citation

> Nutter B, Garbett S, Obregon S, Obadia T, Lehr M, High B, Lane S,
> Beasley W, Gray W, Kennedy N, Hsi-Nien T, Horner J, Stephens J, Beck C,
> Johnson B, Chase P, Tobias P (2026). *redcapAPI: Accessing data from REDCap
> projects using the API*. R package version 2.12.0.
> <https://doi.org/10.5281/zenodo.10564837>

### Current stewardship and project resources

Current public stewardship appears under VUMC Biostatistics / `vubiostat`, with
Shawn Garbett listed as maintainer in current package documentation and the
upstream project README stating that ownership transfer to VUMC Biostatistics is
complete.

Useful current references:

- VUMC Biostatistics redcapAPI project page and abstract by Savannah Obregon,
  Shawn Garbett, and Benjamin Nutter:
  <https://www.vumc.org/biostatistics/node/565>
- Current GitHub repository for the package:
  <https://github.com/vubiostat/redcapAPI>
- Current package site:
  <https://vubiostat.r-universe.dev/redcapAPI>
