---
title: "Getting started with redcapmissing"
vignette: >
  %\VignetteIndexEntry{Getting started with redcapmissing}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`redcapmissing` builds branching-aware missingness reports for REDCap record exports.

This package is built on top of `redcapAPI`. In real workflows, callers create
an `rcon` object with `redcapAPI::redcapConnection()`, export records with an
explicit `redcapAPI::exportRecordsTyped()` cast strategy, and then pass both
objects into `find_missing()`.

The examples below are intentionally synthetic so the vignette remains runnable
without live REDCap credentials or access to a specific REDCap project.

## Why the package uses multiple missingness scopes

REDCap exports are sparse by design. A field can be missing because a value is
blank, but it can also be missing because REDCap exported no row at all for the
needed event or repeat context. The package therefore separates:

- `event_absent`: the expected longitudinal event row is missing
- `repeat_absent`: the expected repeat-instance row is missing
- `form_blank`: the row exists, but every form field is blank or unchecked
- `any_field_missing`: the row context is evaluable, but at least one expected
  field is blank
- `field`: the row exists and the form is started, but an expected field value is missing

This separation keeps the report aligned with how REDCap actually exports data.

## A minimal synthetic REDCap example

```{r}
library(redcapmissing)

meta_row <- function(
  field_name,
  form_name,
  field_type = "text",
  field_label = field_name,
  choices = "",
  branching = "",
  required = ""
) {
  tibble::tibble(
    field_name = field_name,
    form_name = form_name,
    field_type = field_type,
    field_label = field_label,
    select_choices_or_calculations = choices,
    text_validation_type_or_show_slider_number = "",
    branching_logic = branching,
    required_field = required
  )
}

fake_rcon <- function(metadata, mapping = NULL, repeat_instrument_event = NULL) {
  list(
    metadata = function() metadata,
    mapping = function() mapping,
    mappings = function() mapping,
    repeatInstrumentEvent = function() repeat_instrument_event
  )
}
```

## Real `redcapAPI` workflow

In routine use, the package expects a `redcapAPI` connection and a typed export
that keeps coded branching-logic fields in REDCap code space while leaving
checkbox and system fields raw:

```{r, eval = FALSE}
rcon <- redcapAPI::redcapConnection(
  url = Sys.getenv("REDCAP_API_URL"),
  token = Sys.getenv("REDCAP_API_TOKEN")
)

records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    truefalse = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)

report <- find_missing(
  data = records,
  rcon = rcon,
  form = "baseline_form"
)
```

If your project uses additional coded multiple-choice field types, extend this
same pattern so those fields are also kept in code space for downstream
branching-logic-aware missingness checks.

## Required and branching-aware field checks

```{r}
metadata <- dplyr::bind_rows(
  meta_row("record_id", "baseline_form", required = "y"),
  meta_row("branch_flag", "baseline_form", field_type = "yesno", required = "y"),
  meta_row("required_note", "baseline_form", required = "y"),
  meta_row("conditional_note", "baseline_form", branching = "[branch_flag] = '1'", required = "y")
)

records <- tibble::tibble(
  record_id = c("r1", "r2"),
  branch_flag = c("1", "0"),
  required_note = c("entered", "entered"),
  conditional_note = c("", "")
)

report <- find_missing(
  data = records,
  rcon = fake_rcon(metadata),
  form = "baseline_form"
)

report$missing[, c("record_id", "field_name", "missing_scope")]
```

## Restricting a form to selected events

Some REDCap projects use many events of the same general type. In those cases,
`events` lets you keep the assessment focused on the specific events that
should count for the current review.

```{r, eval = FALSE}
followup_report <- find_missing(
  data = records,
  rcon = rcon,
  form = "patient_status",
  events = c(
    "follow_up_1_arm_1",
    "follow_up_2_arm_1",
    "follow_up_3_arm_1"
  )
)
```

If `events` is omitted, the function defaults to all offered events for
that form. If the form is only offered on one event, the argument is ignored.

## Forms that are regular on some events and repeating on others

Some REDCap projects offer the same form in mixed ways across events. For
example, a form might be:

- a regular form on `event_a`
- repeating on `event_b`
- a regular form again on `event_c`

This matters because the package should not activate repeat-instance logic just
because the form repeats somewhere in the project. It should only activate that
logic for the requested events where the form actually repeats.

Operationally, that means:

- if you request only non-repeating events for that form, the function uses the
  standard `field`, `any_field_missing`, `form_blank`, and `event_absent`
  scopes
- if you request repeating events for that form, the function uses `field`,
  `any_field_missing`, `form_blank`, and `repeat_absent` for those repeating
  contexts
- if you request a mixture of repeating and non-repeating events, the function
  applies the appropriate scope model per event
- if `instances` is omitted and at least one requested event is repeating
  for that form, the function assumes `1L` only for those requested repeating
  contexts

```{r, eval = FALSE}
mixed_report <- find_missing(
  data = records,
  rcon = rcon,
  form = "patient_status",
  events = c(
    "event_a_arm_1",
    "event_b_arm_1",
    "event_c_arm_1"
  )
)
```

In that example, `event_a_arm_1` and `event_c_arm_1` would be checked as regular
form events, while `event_b_arm_1` would be checked with repeat-instance logic.

## Repeat expectations

`instances` exists because REDCap does not export nonexistent repeat rows.
If instance 2 should exist but was never created, there is no row in the export
for the package to inspect unless it first creates the expected repeat contexts.

```{r}
repeat_metadata <- dplyr::bind_rows(
  meta_row("record_id", "screen_form", required = "y"),
  meta_row("screen_started", "screen_form", required = "y"),
  meta_row("repeat_started", "repeat_form", required = "y"),
  meta_row("repeat_value", "repeat_form", required = "y")
)

mapping <- tibble::tibble(
  arm_num = c(1, 1),
  unique_event_name = c("baseline_event", "baseline_event"),
  form = c("screen_form", "repeat_form")
)

repeat_map <- tibble::tibble(
  event_name = "baseline_event",
  form_name = "repeat_form",
  custom_form_label = ""
)

repeat_records <- tibble::tibble(
  record_id = c("r1", "r2", "r2"),
  redcap_event_name = c("baseline_event", "baseline_event", "baseline_event"),
  redcap_repeat_instrument = c("repeat_form", "repeat_form", "repeat_form"),
  redcap_repeat_instance = c("1", "1", "2"),
  repeat_started = c("yes", "yes", "yes"),
  repeat_value = c("10", "20", "30")
)

repeat_report <- find_missing(
  data = repeat_records,
  rcon = fake_rcon(repeat_metadata, mapping = mapping, repeat_instrument_event = repeat_map),
  form = "repeat_form",
  instances = 2L
)

repeat_report$repeat_missing[, c("record_id", "redcap_repeat_instance", "missing_scope")]
```

## Inspecting the return object

The report object is intentionally multi-part. In practice, these are the most
useful components to inspect first:

- `report$missing`: all failed rows extracted from the interrogated agent
- `report$form_missing`: the form-level failures for wholly blank form contexts
- `report$event_missing`: the failures for expected event rows that were absent
- `report$repeat_missing`: the failures for expected repeat rows that were absent
- `report$any_field_missing`: the patient-context roll-up failures for
  evaluable contexts with at least one missing expected field
- `report$expected`: the field-level expected rows that were actually assessed
- `report$agent`: the interrogated `pointblank` agent itself

This design keeps the correction queue (`report$missing`) separate from the
full validation context and from the scope-specific helper tables.

## Tidy summary tables

`tidy(report)` returns a focused validation summary tibble with one row per
validation step and REDCap context, including `validation_context` and the
REDCap event/repeat columns used for stratified denominators. If optional
formatting packages are installed, you can also format that validation summary
for reporting workflows:

```{r, eval = FALSE}
tidy_tbl <- tidy(report)
ft <- flex(report)
html <- flex_html(ft)
```

## Citation and acknowledgement

`redcapmissing` depends heavily on `redcapAPI`. Please cite `redcapAPI` when
`redcapmissing` is part of a REDCap workflow.

### Foundational package citation

> Nutter B, Garbett S, Obregon S, Obadia T, Lehr M, High B, Lane S, Beasley W,
> Gray W, Kennedy N, Hsi-Nien T, Horner J, Stephens J, Beck C, Johnson B,
> Chase P, Tobias P (2026). *redcapAPI: Accessing data from REDCap projects
> using the API*. R package version 2.12.0.
> <https://doi.org/10.5281/zenodo.10564837>

### Current stewardship and project resources

Current public stewardship appears under VUMC Biostatistics / `vubiostat`, with
Shawn Garbett listed as maintainer in current package documentation and the
upstream project README stating that ownership transfer to VUMC Biostatistics is
complete.

Useful current references:

- VUMC Biostatistics redcapAPI project page and abstract by Savannah Obregon,
  Shawn Garbett, and Benjamin Nutter:
  <https://www.vumc.org/biostatistics/node/565>
- Current GitHub repository for the package:
  <https://github.com/vubiostat/redcapAPI>
- Current package site:
  <https://vubiostat.r-universe.dev/redcapAPI>
