---
title: "Getting started with redcapmissing"
vignette: >
  %\VignetteIndexEntry{Getting started with redcapmissing}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  default_output_hook(gsub("[[:blank:]]+(?=\\r?\\n|$)", "", x, perl = TRUE), options)
})
library(redcapmissing)
```

`redcapmissing` builds branching-aware missingness reports for REDCap record
exports. In real workflows, callers create a
`redcapAPI::redcapConnection()`, export records with
`redcapAPI::exportRecordsTyped()`, and pass both objects into `find_missing()`.

The examples below are synthetic so the vignette can run without live REDCap
credentials.

## Validation registry

The package validation model is available through `registry()`:

```{r}
registry()
```

The four checks run in registry order. Report rows use `event:form` for
non-repeating contexts and `event:form:instance` for repeat-instance contexts;
any failed check removes that context from downstream assessment. See
`?registry` for the complete taxonomy and column definitions.

## Synthetic REDCap helpers

```{r}
meta_row <- function(
  field_name,
  form_name,
  field_type = "text",
  field_label = field_name,
  choices = "",
  branching = "",
  required = ""
) {
  tibble::tibble(
    field_name = field_name,
    form_name = form_name,
    field_type = field_type,
    field_label = field_label,
    select_choices_or_calculations = choices,
    text_validation_type_or_show_slider_number = "",
    branching_logic = branching,
    required_field = required
  )
}

fake_rcon <- function(metadata, mapping = NULL, repeat_instrument_event = NULL) {
  forms <- unique(as.character(metadata$form_name))
  instruments <- tibble::tibble(
    instrument_name = forms,
    instrument_label = paste(forms, "label")
  )

  list(
    metadata = function() metadata,
    instruments = function() instruments,
    mapping = function() mapping,
    mappings = function() mapping,
    repeatInstrumentEvent = function() repeat_instrument_event
  )
}
```

## Real `redcapAPI` workflow

For real REDCap projects, keep branching-logic fields in REDCap code space and
leave checkbox and system fields raw:

```{r, eval = FALSE}
rcon <- redcapAPI::redcapConnection(
  url = Sys.getenv("REDCAP_API_URL"),
  token = Sys.getenv("REDCAP_API_TOKEN")
)

records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    truefalse = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)

report <- find_missing(
  data = records,
  rcon = rcon,
  forms = "baseline_form"
)
```

## Branching-aware field checks

```{r}
metadata <- dplyr::bind_rows(
  meta_row("record_id", "baseline_form", required = "y"),
  meta_row("branch_flag", "baseline_form", field_type = "yesno", required = "y"),
  meta_row("required_note", "baseline_form", required = "y"),
  meta_row(
    "conditional_note",
    "baseline_form",
    branching = "[branch_flag] = '1'",
    required = "y"
  )
)

records <- tibble::tibble(
  record_id = c("r1", "r2"),
  branch_flag = c("1", "0"),
  required_note = c("entered", "entered"),
  conditional_note = c("", "")
)

report <- find_missing(
  data = records,
  rcon = fake_rcon(metadata),
  forms = "baseline_form"
)

get_missing(
  report,
  validation_check = "field-complete"
)[, c(
  "record_id",
  "validation_context",
  "form",
  "validation_check",
  "field_name"
)]
```

Record `r1` is branch-open for `conditional_note`, so the missing value is a
`field-complete` failure. Record `r2` is branch-closed, so the field is not in
the assessment pool.

The `url` column from `get_missing(report)` is a raw REDCap Data Entry URL for
each failed record/form context when `rcon` includes the required instance,
version, and project metadata, plus an event ID for longitudinal projects. It
is `NA` for connection-like objects that do not provide that metadata,
including the synthetic helper used here.

## Event-row gating

REDCap exports may omit entire event rows. `event-row-started` catches that
before form and field checks run downstream.

```{r}
status_metadata <- dplyr::bind_rows(
  meta_row("record_id", "status_form", required = "y"),
  meta_row("status_started", "status_form", required = "y"),
  meta_row("status_value", "status_form", required = "y")
)

mapping <- tibble::tibble(
  arm_num = c(1, 1, 1),
  unique_event_name = c("event_1_arm_1", "event_2_arm_1", "event_3_arm_1"),
  form = c("status_form", "status_form", "status_form")
)

event_records <- tibble::tibble(
  record_id = c("r1", "r1", "r2", "r2"),
  redcap_event_name = c(
    "event_1_arm_1",
    "event_2_arm_1",
    "event_1_arm_1",
    "event_2_arm_1"
  ),
  status_started = c("yes", "yes", "yes", "yes"),
  status_value = c("entered", "", "entered", "entered")
)

event_report <- find_missing(
  data = event_records,
  rcon = fake_rcon(status_metadata, mapping = mapping),
  forms = "status_form"
)

get_summary(event_report)[
  ,
  c("validation_level", "validation_check", "redcap_event_name", "assessed", "failed")
]
```

`event_3_arm_1` fails `event-row-started`, so it is not assessed by
`form-started` or `field-complete`.

## Event-specific record eligibility

Use `records` when records become eligible for validation on different REDCap
events, forms, or repeat instances at different times. The argument is a named
list keyed by raw `redcap_event_name`. An event-level vector limits every final
requested form and selected repeat instance at that event. A nested form vector
limits one event/form context. A nested instance list limits one
event/form/repeat-instance context. Omitted event, form, and instance entries
use the default eligibility implied by `data`, `forms`, `events`, `instances`,
and `ignore_ids`. Empty, missing, `NULL`, and blank-only record values are
errors. To remove an event from assessment entirely, exclude it with `events`
rather than using an empty `records` entry.

Every report stores `spec$record_eligibility` with one row for each final
assessed record/event/form/repeat-instance context, even when `records` is
omitted. If filtering leaves no assessable record IDs, `find_missing()` stops.
Explicit `records` entries are the exception because they can declare expected
contexts whose absent data rows should fail a row-started check.

```{r}
staged_metadata <- dplyr::bind_rows(
  meta_row("record_id", "demographics", required = "y"),
  meta_row("demographics_started", "demographics", required = "y"),
  meta_row("demographics_value", "demographics", required = "y"),
  meta_row("surgery_started", "surgery", required = "y"),
  meta_row("surgery_value", "surgery", required = "y")
)

staged_mapping <- tibble::tibble(
  arm_num = c(1, 1, 1),
  unique_event_name = c("event_a", "event_b", "event_c"),
  form = c("demographics", "surgery", "surgery")
)

staged_records <- tibble::tibble(
  record_id = c("record_a", "record_a", "record_b", "record_b", "record_b"),
  redcap_event_name = c("event_a", "event_b", "event_a", "event_b", "event_c"),
  demographics_started = c("yes", "", "yes", "", ""),
  demographics_value = c("entered", "", "entered", "", ""),
  surgery_started = c("", "yes", "", "yes", "yes"),
  surgery_value = c("", "entered", "", "entered", "")
)

staged_report <- find_missing(
  data = staged_records,
  rcon = fake_rcon(staged_metadata, mapping = staged_mapping),
  forms = c("surgery", "demographics"),
  records = list(
    event_a = c("record_a", "record_b"),
    event_b = c("record_a", "record_b"),
    event_c = list(
      surgery = "record_b"
    )
  )
)

staged_report$spec$record_eligibility

staged_summary <- get_summary(staged_report)
staged_summary[
  staged_summary$validation_check == "event-row-started",
  c("form", "redcap_event_name", "assessed", "failed")
]
```

Record `record_a` is assessed for `surgery` on `event_b`, but not on
`event_c`. Record `record_b` is assessed on `event_c`, where its blank surgery
value fails downstream field completeness.

If another requested event offered `demographics` and was not named in
`records`, the `records` argument would not limit that event. The package would
consider every applicable non-ignored record in `data` for that event, then use
`event-row-started` to decide whether each record actually has an event row.

## Repeat expectations

REDCap does not export nonexistent repeat rows. Use `instances` to declare the
repeat instances expected for the report.

```{r}
repeat_metadata <- dplyr::bind_rows(
  meta_row("record_id", "screen_form", required = "y"),
  meta_row("screen_started", "screen_form", required = "y"),
  meta_row("repeat_started", "repeat_form", required = "y"),
  meta_row("repeat_value", "repeat_form", required = "y")
)

repeat_mapping <- tibble::tibble(
  arm_num = c(1, 1),
  unique_event_name = c("baseline_event", "baseline_event"),
  form = c("screen_form", "repeat_form")
)

repeat_map <- tibble::tibble(
  event_name = "baseline_event",
  form_name = "repeat_form",
  custom_form_label = ""
)

repeat_records <- tibble::tibble(
  record_id = c("r1", "r2", "r2"),
  redcap_event_name = c("baseline_event", "baseline_event", "baseline_event"),
  redcap_repeat_instrument = c("repeat_form", "repeat_form", "repeat_form"),
  redcap_repeat_instance = c("1", "1", "2"),
  repeat_started = c("yes", "yes", "yes"),
  repeat_value = c("10", "20", "30")
)

repeat_report <- find_missing(
  data = repeat_records,
  rcon = fake_rcon(
    repeat_metadata,
    mapping = repeat_mapping,
    repeat_instrument_event = repeat_map
  ),
  forms = "repeat_form",
  instances = 2L
)

get_missing(
  repeat_report,
  validation_check = "instance-row-started"
)[, c(
  "record_id",
  "redcap_event_name",
  "redcap_repeat_instrument",
  "redcap_repeat_instance",
  "validation_context",
  "form",
  "validation_check"
)]
```

The missing instance `2` for record `r1` fails `instance-row-started` and is
removed from downstream form and field assessment.

## Mixed regular and repeating contexts

Some REDCap projects offer the same form as regular on one event and repeating
on another. The package activates event/form checks per context:

- regular event contexts use `event-row-started`
- repeating contexts use `instance-row-started`

If `events` excludes the repeating event, `instances` is ignored for that form.
If a requested repeating context omits `instances`, the package assumes
instance `1` and warns once.

## Inspecting results

Use `get_summary()` to review validation counts and rates. Use `get_missing()`
to identify the failed record or field contexts that need follow-up. Both
accessors can focus the report by validation check, event, or form.

```{r}
summary_result <- get_summary(report)
summary_result

missing_result <- get_missing(
  report,
  validation_check = "field-complete"
)
missing_result[, c("record_id", "form", "field_name", "field_label")]
```

See `?get_summary` and `?get_missing` for their complete filtering and return
contracts.

For presentation, `flexify()` turns either accessor result into a flextable,
while `flex_event_forms()` provides a compact event/form overview of due,
started, and incomplete form opportunities:

```{r, eval = FALSE}
summary_ft <- flexify(summary_result)
missing_ft <- flexify(missing_result)
event_form_ft <- flex_event_forms(event_report)
```

Use the accessor tables for detailed review and `flex_event_forms()` for a
high-level reporting view. See `?flexify` for compatible input shapes and
formatting behavior, and `?flex_event_forms` for the complete metric and
threshold definitions.
