---
title: "Creating and running an assessment plan"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Creating and running an assessment plan}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(redcapmissing)
```

`redcapmissing` separates target selection from assessment:

1. `all_instruments()`, `build_extended_schedule()`, and
   `build_explicit_schedule()` help declare project instrument scope and
   schedules.
2. `plan_from_data()` or `plan_explicit()` freezes the record, event,
   instrument, and repeat instance crossings stored in `assessible_targets`.
3. `run_plan()` evaluates the physical rows, instruments, and fields for those
   frozen targets.

This vignette explains the complete workflow from record export through result
inspection.

## Before you begin

The examples assume that the caller has supplied these objects:

| Object | Required for | Contents |
|---|---|---|
| `rcon` | Planning helpers, both constructors, and `run_plan()` | A `redcapAPI` object inheriting from `redcapApiConnection` or `redcapOfflineConnection` that represents the project used for `data`. |
| `instruments` | The extended schedule builder and both constructors | A character vector of raw REDCap instrument names, optionally returned by `all_instruments(rcon)`. |
| `records` | Both constructors and `run_plan()` | Records exported from the project represented by `rcon`. |
| `extended_schedule` | Optional use of `plan_from_data()` | A table with the exact three column schema described below. |
| `explicit_schedule` | `plan_explicit()` | A table with the exact four column schema described below. |
| `verified` and `verified_user` | Optional use of `run_plan()` | A verification table and the exact username whose rows may apply. |

`redcapApiConnection` objects are returned by `redcapAPI::redcapConnection()`;
`redcapOfflineConnection` objects are returned by
`redcapAPI::offlineConnection()` or `redcapAPI::readPreservedProject()`. For an
online connection, use `redcapAPI::unlockREDCap()` with its encrypted keyring
when available, or obtain a token at run time. Keep the token only in memory and
exclude it from R source, rendered documents, logs, and saved workspaces.

## Export records

Export typed records while retaining REDCap codes for fields used by branching
logic. Keep checkbox and REDCap system columns in their exported form.

```{r export, eval = FALSE}
records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    truefalse = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)
```

The constructors discover the project record ID field from `rcon`. The field
may have any valid REDCap field name. Planning treats every exported physical
row as observed, including rows whose response values are all missing. Export
records before constructing a plan: neither constructor exports records
through `rcon`.

## Planning functions

| Function | Arguments | Result |
|---|---|---|
| `all_instruments()` | `rcon` | The complete ordered raw instrument inventory. |
| `build_extended_schedule()` | `rcon`, `instruments`, and optional `n_repeat_instances` | A three-column extension containing every allowable crossing for the requested instruments. |
| `build_explicit_schedule()` | pipe-first `data`, `rcon`, and `explicit_spec` | A four-column explicit schedule crossing unique project record IDs with selected allowable crossings. |
| `plan_from_data()` | `data`, `rcon`, `instruments`, and optional `extended_schedule` | Frozen targets observed in `data`, plus those added by the extension. |
| `plan_explicit()` | `data`, `rcon`, and required `explicit_schedule` | Frozen targets and instrument scope named in `explicit_schedule`. |

For both constructors:

- `data` is a data frame or tibble from the project represented by `rcon`.
- `rcon` supplies project identity, metadata, instruments, repeat settings,
  and the applicable arms, events, and instrument event mappings.

For `plan_from_data()`, `instruments` is a nonempty character vector of unique,
nonmissing, nonblank, unpadded raw instrument names. Its order controls target
ordering. `plan_explicit()` has no separate `instruments` argument: its complete
instrument scope is derived from normalized `explicit_schedule` rows.

Caller-supplied `data` is the only physical-row source for both constructors.
The constructors read project structure from `rcon` but never use it to export
records.

## Plan from observed data

`plan_from_data()` applies this rule:

```text
assessible_targets =
  allowable crossings described by rcon
  INTERSECT
  (
    crossings observed in data
    UNION
    crossings in extended_schedule
  )
```

```{r observed_plan, eval = FALSE}
observed_plan <- plan_from_data(
  data = records,
  rcon = rcon,
  instruments = instruments
)
```

`data` must contain at least one physical row. Instrument response columns are
optional during planning.

`extended_schedule = NULL` selects observed planning. A zero row extension
with the complete schema has the same result.

## Add crossings with `extended_schedule`

Use `all_instruments()` when the plan should declare the complete project
inventory, then pass all or a subset of that vector to the builder:

```{r build_extension, eval = FALSE}
instruments <- all_instruments(rcon)
extension_instruments <- c("followup", "diary")
extended_schedule <- build_extended_schedule(
  rcon = rcon,
  instruments = extension_instruments,
  n_repeat_instances = 3L
)
```

The builder requires a nonempty vector of unique project instrument names.
`NULL` does not mean all instruments; call `all_instruments(rcon)` explicitly.

The result has exactly `instrument`, `redcap_event_name`, and
`repeat_instance`, with character, character, and integer storage.

Classic project rows use the native eventless crossing represented by
`redcap_event_name = NA_character_`. A classic repeating instrument receives
instances `1` through `n_repeat_instances`. Longitudinal rows always use a
concrete event designated for that instrument. A valid longitudinal instrument
designated to no event contributes no row and emits one
`redcapmissing_warning_undesignated_extension` warning; the builder never
creates a longitudinal crossing with a missing event.

If none of the requested longitudinal instruments is designated to an event,
the builder warns and returns a correctly typed zero-row schedule. To request
no extension, omit `extended_schedule` from `plan_from_data()`.

The extension has exactly these columns in this order:

| Column | Accepted storage and values | Accepted missing values | Stored in the plan |
|---|---|---|---|
| `instrument` | Character or factor containing a value from `instruments`; values are nonblank and unpadded. | None. | Character. |
| `redcap_event_name` | Character or factor containing a raw event name in a longitudinal project. | In a classic project, character or factor blank values and typed `NA` values are accepted. Accepted noncharacter storage is a nonempty atomic vector containing only typed `NA` values. `Date`, `POSIXt`, and `NaN` are rejected even when the complete column is missing. | Character; `NA_character_` for a classic project. |
| `repeat_instance` | Positive integer, finite whole number double, or character digits matching `[1-9][0-9]*` within R integer range when the scheduled event or instrument repeats. | When neither the scheduled event nor instrument repeats, numeric `NA`, character `NA` or blank, or a nonempty logical or factor column whose every value is `NA` is accepted. `Date`, `POSIXt`, and `NaN` are rejected even when the complete column is missing. | Integer; `NA_integer_` when neither the scheduled event nor instrument repeats. |

A zero row extension uses character or factor storage for `instrument` and
`redcap_event_name`, and character, integer, or double storage for
`repeat_instance`. Column order is checked even when the table has no rows.
Extra columns, duplicate rows after value conversion, unknown names, and
crossings disallowed by `rcon` produce an error.

```{r extension, eval = FALSE}
extended_plan <- plan_from_data(
  data = records,
  rcon = rcon,
  instruments = instruments,
  extended_schedule = extended_schedule
)
```

Each row adds its exact instrument, event, and repeat instance crossing for
records observed in the applicable arm. In a classic project, it adds the
crossing for every observed record. An arm with no observed records produces
one `redcapmissing_warning_empty_arm_extension` warning and contributes no
targets. Each row contributes only the instance named in that row.

An omitted crossing receives the observed planning rule. Any allowable
crossing observed in `data` remains in `assessible_targets`.

## Plan from `explicit_schedule`

`plan_explicit()` applies this rule:

```text
assessible_targets =
  allowable crossings described by rcon
  INTERSECT
crossings in explicit_schedule
```

Use `build_explicit_schedule()` to create that exact schedule from cohort data
and selected project mappings. It discovers the record-ID field from `rcon`;
the input must retain that exact field name, such as `participant_id`. The
output always uses canonical `record_id` for `plan_explicit()`.

```{r build_explicit, eval = FALSE}
explicit_spec_1 <- rcon$mapping() |>
  dplyr::filter(unique_event_name %in% desired_events_1)

explicit_schedule_1 <- data.frame(
  participant_id = c("001", "002", "003")
) |>
  build_explicit_schedule(rcon, explicit_spec_1)

explicit_spec_2 <- rcon$mapping() |>
  dplyr::filter(
    unique_event_name %in% desired_events_2,
    form == "diary"
  ) |>
  dplyr::mutate(repeat_instance = 2L)

explicit_schedule_2 <- data.frame(
  participant_id = c("008", "009", "010")
) |>
  build_explicit_schedule(rcon, explicit_spec_2)

explicit_schedule <- dplyr::bind_rows(
  explicit_schedule_1,
  explicit_schedule_2
)
```

`explicit_spec` contains exactly one event alias (`unique_event_name` or
`redcap_event_name`) and one instrument alias (`form` or `instrument`). Extra
columns such as `arm_num` are ignored, so a filtered `rcon$mapping()` result can
be passed directly. If `repeat_instance` is absent, the builder supplies
`NA_integer_`; repeating events and instruments instead require supplied
positive instances. Specifications are validated against current project
structure before expansion. Record IDs are unique in first-appearance order,
and specification rows retain their order within every record.

Schedules made independently can be combined with `dplyr::bind_rows()`.
Disjoint cohorts and overlapping records with different crossings are valid.
An exact duplicate target across pieces remains an error when the combined
schedule is passed to `plan_explicit()`.

The schedule has exactly these columns in this order:

| Column | Accepted storage and values | Accepted missing values | Stored in the plan |
|---|---|---|---|
| `record_id` | Character, factor, integer, or finite double; values are nonblank and unpadded. Character leading zeros are preserved. | None. | Character. |
| `instrument` | Character or factor containing a project raw instrument name; values are nonblank and unpadded. | None. | Character. |
| `redcap_event_name` | Character or factor containing a raw event name in a longitudinal project. | In a classic project, character or factor blank values and typed `NA` values are accepted. Accepted noncharacter storage is a nonempty atomic vector containing only typed `NA` values. `Date`, `POSIXt`, and `NaN` are rejected even when the complete column is missing. | Character; `NA_character_` for a classic project. |
| `repeat_instance` | Positive integer, finite whole number double, or character digits matching `[1-9][0-9]*` within R integer range when the scheduled event or instrument repeats. | When neither the scheduled event nor instrument repeats, numeric `NA`, character `NA` or blank, or a nonempty logical or factor column whose every value is `NA` is accepted. `Date`, `POSIXt`, and `NaN` are rejected even when the complete column is missing. | Integer; `NA_integer_` when neither the scheduled event nor instrument repeats. |

A zero row explicit schedule uses the same column storage described for the
extension, with character, factor, integer, or double storage for `record_id`.
It creates a valid plan with no targets. `explicit_schedule = NULL` and an
omitted `explicit_schedule` produce an error.

```{r explicit_plan, eval = FALSE}
explicit_plan <- plan_explicit(
  data = records,
  rcon = rcon,
  explicit_schedule = explicit_schedule
)
```

Each schedule row creates its exact target. A record ID absent from `data`
remains in the plan. During `run_plan()`, an absent longitudinal event fails
`event-row-started`; a present event with an absent repeat instance fails
`repeat-instance-row-started`; and an absent classic target with
`repeat_instance = NA_integer_` has both row checks marked `"not applicable"`
and fails `instrument-started`.
For a longitudinal project, a record already observed in one arm must be
scheduled within that arm.
An instrument with no explicit schedule row is outside the explicit plan's
scope and produces no warning. A correctly typed zero-row schedule creates a
plan with `instruments = character()` and no targets.

The absence rules differ:

| Input | Row absent | Table is `NULL` | Complete table has zero rows |
|---|---|---|---|
| `extended_schedule` | Use the observed planning rule for that crossing. | Use observed planning. | Use observed planning. |
| `explicit_schedule` | Exclude that exact target. | Error. | Create a plan with no targets. |

## Structural columns in `data`

Every supplied column must be an atomic vector with no dimensions. Structural
values are converted as follows:

| Column or dimension | Accepted storage and values | Missing value rule | Value used by the package |
|---|---|---|---|
| Project record ID field | Character, factor, integer, or finite double; values are nonblank and unpadded. | Missing values, `Date`, `POSIXt`, `NaN`, and infinity produce an error. | Character; character leading zeros are preserved. |
| `redcap_event_name` in a longitudinal project | Character or factor raw event names. | Missing and blank values produce an error. | Character. |
| `redcap_event_name` in a classic project | The column may be absent. If present, it uses character or factor storage, or a nonempty atomic column containing only typed `NA` values. | Every value is blank or missing. `Date`, `POSIXt`, and `NaN` are rejected even when every value in the column is missing. | `NA_character_`. |
| `redcap_repeat_instrument` | Character or factor raw instrument names where a repeating instrument applies. Accepted noncharacter storage is a nonempty atomic vector containing only typed `NA` values. | Blank or missing applies to rows at a repeating event and rows where neither the event nor instrument repeats. `Date`, `POSIXt`, and `NaN` are rejected even when every value in the column is missing. | Character or `NA_character_`. |
| `redcap_repeat_instance` | Positive integer, finite whole number double, or character digits matching `[1-9][0-9]*` within R integer range when the event or instrument repeats. | When neither the event nor instrument repeats, numeric `NA`, character `NA` or blank, or a nonempty logical or factor column whose every value is `NA` is accepted. `Date`, `POSIXt`, and `NaN` are rejected even when every value in the column is missing. | Integer or `NA_integer_`. |

When the project contains a repeating event or repeating instrument, `data`
must include `redcap_repeat_instrument` and `redcap_repeat_instance`. Supplying
one of these columns also requires the other.

The row key is the converted record ID, event, repeat instrument, and repeat
instance. Each key occurs once. A collision created during value conversion
produces a duplicate key error.

`plan_from_data()` requires at least one row. `plan_explicit()` and
`run_plan()` accept a zero row data frame with every structural column required
by the project.

## Classic, longitudinal, and repeating projects

Classic projects store `NA_character_` for `redcap_event_name` because they
have no event dimension. An extension applies to all observed records. A
classic repeating instrument uses positive instances at that eventless
crossing; neither case is an undesignated instrument or produces that warning.

Longitudinal projects require one raw event name on every physical row.
Extensions apply to records observed in the arm that contains the scheduled
event. An instrument mapped at only some events is assessible only at those
events. A project instrument designated to no longitudinal event remains part
of the project inventory but contributes no observed, extended, or explicit
target unless REDCap later designates it to an event.

A row where neither its event nor its instrument repeats stores
`NA_character_` in `redcap_repeat_instrument` and `NA_integer_` in
`redcap_repeat_instance`. A row at a repeating event stores `NA_character_`
for the repeat instrument and a positive instance. A row for a repeating
instrument stores the raw instrument name and a positive instance.

An instrument may have no repeat at one event and repeat at another. The
schedule row must match the repeat setting for that event.

## Plan contents

Both constructors return a `redcapmissing_plan` with these components:

| Component | Storage and values |
|---|---|
| `schema_version` | Integer scalar `1L`. |
| `construction` | Character scalar `"from_data"` or `"explicit"`. |
| `instruments` | Character vector in the requested order. |
| `assessible_targets` | Tibble with the six columns below. |
| `project` | List containing `project_id`, `record_id_field`, `longitudinal`, `event_labels`, and `instrument_labels`. |
| `structure_fingerprint` | Character scalar containing a 64 character lowercase hexadecimal SHA-256 digest. |

`project$project_id` and `project$record_id_field` are character scalars.
`project$longitudinal` is one logical value. `project$event_labels` and
`project$instrument_labels` are named character vectors; the event vector is
empty for a classic project.

`assessible_targets` contains:

| Column | Storage and values |
|---|---|
| `record_id` | Character record ID. |
| `instrument` | Character raw instrument name from `instruments`. |
| `redcap_event_name` | Character raw event name, or `NA_character_` for a classic project. |
| `repeat_instrument` | Character raw instrument name for a repeating instrument target; `NA_character_` for a repeating event target or when neither the event nor instrument repeats. |
| `repeat_instance` | Positive integer when the event or instrument repeats; `NA_integer_` when neither repeats. |
| `target_source` | `"observed"`, `"extended"`, `"observed+extended"`, or `"explicit"` as allowed by `construction`. |

Targets are unique on the first five columns. Their order follows
`instruments`, REDCap event order, record ID, repeat kind, and repeat instance.
The fingerprint represents project identity, metadata, instruments, arms,
events, instrument event mappings, and repeat settings independent of source
table row order and the current `scipen` and `OutDec` option values. The stored
components exclude the input `data` and `rcon`.

## Run a plan

```{r run, eval = FALSE}
report <- run_plan(
  plan = extended_plan,
  data = records,
  rcon = rcon
)
```

| Argument | Required | Default | Accepted values | Effect |
|---|---:|---|---|---|
| `plan` | Yes | None | A validated `redcapmissing_plan`. | Supplies the stored targets, instrument scope, `target_source` provenance, project data, and fingerprint. |
| `data` | Yes | None | A data frame or tibble with structural columns, response columns needed by frozen target instruments, and branching dependencies for those responses. A complete zero row table is valid. | Supplies the rows and response values assessed during this call. |
| `rcon` | Yes | None | The connection or supported object for the same project structure represented by `plan`. | Supplies current project identity and structure. |
| `required_fields` | No | `TRUE` | One nonmissing logical value. | `TRUE` retains fields marked required before later removals; `FALSE` retains required and optional fields. |
| `ignore_fields` | No | `NULL` | `NULL`, `character(0)`, or unique nonmissing, nonblank, unpadded raw metadata field names that remain after the `required_fields` filter and `exclude_types` removal. Unknown names and names removed by an earlier step produce an error. | Removes those names from `field-complete`. |
| `exclude_types` | No | `"descriptive"` | `NULL`, `character(0)`, or unique nonmissing, nonblank, unpadded metadata field types present after the `required_fields` filter. An explicitly supplied unused type produces an error. The omitted default removes `descriptive` fields when present and is accepted when none remain. | Removes those types from `field-complete`. |
| `verified` | No | `NULL` | `NULL`, a complete zero row verification table, or a table with the nine columns described below. | Supplies verification evidence for failed `field-complete` results. |
| `verified_user` | No | `NULL` | `NULL` or one nonmissing, nonblank, unpadded character username. | Selects rows whose `username` exactly matches this value, including case. |
| `details` | No | `FALSE` | One nonmissing logical value. | `TRUE` stores check and assessed field rows in `details`; `FALSE` stores `NULL`. |
| `progress` | No | `interactive()` | One nonmissing logical value. | Controls progress reporting for the twelve stages. |

The current `data` may contain rows exported after plan construction when
`rcon` still describes the same project structure. The plan continues to
supply its stored targets. Rows outside those targets are ignored. For an
absent stored target, a missing longitudinal event fails `event-row-started`, a
missing repeat instance after a present event fails
`repeat-instance-row-started`, and a missing classic target with
`repeat_instance = NA_integer_` fails `instrument-started` after both row
checks receive `"not applicable"`.
Runtime field preflight is target-scoped. It derives instrument scope from
`plan$assessible_targets`, then requires detection fields, retained assessment
fields, checkbox child columns, branching dependencies for those fields, and
the structural columns needed to match targets. A selected plan instrument
with no frozen target imposes no exclusive response-column requirement. A
field on a nontarget instrument remains required when target branching logic
references it.

### Execution stages

`diagnostics` records these stages in order:

| Stage | Operation |
|---:|---|
| 1 | `"Validate plan, data, and rcon"` |
| 2 | `"Validate and normalize verification"` |
| 3 | `"Resolve instrument-start fields"` |
| 4 | `"Resolve field-complete fields"` |
| 5 | `"Join assessible_targets to physical rows"` |
| 6 | `"Run event-row-started"` |
| 7 | `"Run repeat-instance-row-started"` |
| 8 | `"Run instrument-started"` |
| 9 | `"Run raw field-complete"` |
| 10 | `"Apply verification"` |
| 11 | `"Aggregate effective results"` |
| 12 | `"Construct the report"` |

### Gates and statuses

Check statuses are `"passed"`, `"failed"`, `"not applicable"`, and
`"not reached"`. Validation levels are `"event:instrument"` and
`"event:instrument:instance"`.

`registry()` documents these validation checks in five columns:
`validation_order` is the check sequence; `validation_level` is the value stored
in report rows; `validation_check` is the exact report and filter code;
`flex_label` is the label used by `flexify()`; and `description` states the pass
condition for an assessed check.

| Target context | `event-row-started` | `repeat-instance-row-started` |
|---|---|---|
| Classic project | `"not applicable"` | Evaluated when `repeat_instance` is positive; `"not applicable"` when it is `NA_integer_`. |
| Longitudinal target with `repeat_instance = NA_integer_` | Passes when any physical row exists for its record and event. | `"not applicable"`. |
| Longitudinal target with a positive `repeat_instance` | Passes when any physical row exists for its record and event. | Evaluated after the event check passes. |

A failed event or repeat instance check sets downstream checks to
`"not reached"`. `instrument-started` then checks its own field set.
`field-complete` runs after `instrument-started` passes.

```{r registry}
registry()
```

## Instrument detection and field selection

`instrument-started` uses every data entry metadata field on each instrument
represented in the frozen targets except:

- the project record ID field;
- fields with type `descriptive`; and
- fields with type `calc`.

A checkbox root expands to its exported child columns. An ordinary detection
field starts the instrument when its response is present. A checkbox root
starts the instrument when at least one exported child column is selected. The
values `"0"`, `"unchecked"`, `"false"`, and `"no"` are unselected regardless
of letter case. A targeted instrument with no usable detection fields produces
an error; a selected instrument with no target does not enter this check.

`required_fields`, `exclude_types`, and `ignore_fields` resolve the
`field-complete` set in this order:

```text
all fields on each targeted instrument
-> retain fields marked required when required_fields = TRUE
-> remove field types named by exclude_types
-> remove field names named by ignore_fields
```

`ignore_fields` contains raw metadata field names. A checkbox uses its root
field name. Unknown, duplicate, missing, blank, padded, and unused names or
types produce an error. Explicit field policies remain strict against the
target-scoped field set.

These three arguments change `field-complete` only. Detection fields remain
available to `instrument-started`. When the policy leaves no fields,
`field-complete` receives:

| Output | Value |
|---|---|
| status | `"not applicable"` |
| reason | `"no fields remain after field policy"` |
| counts | Zero. |
| rates | `NA_real_`. |

## Branching logic and checkboxes

Branching logic is evaluated for each target event and repeat context. Fields
whose logic evaluates to false are excluded from the field count for that
target.

For an unqualified reference to another event:

1. A unique source row whose `redcap_repeat_instance` is missing supplies the
   referenced value.
2. When no such source row exists, one source row with a positive repeat
   instance supplies the value.
3. Multiple source rows with positive repeat instances and no source row with
   a missing repeat instance produce a `redcapmissing_error_project` because
   the reference identifies no single instance.

A target whose selected fields all have false branching logic receives:

| Output | Value |
|---|---|
| `field_complete` | `"not applicable"` |
| `field_applicability_reason` | `"no fields apply after branching logic"` |
| field counts | Zero. |

A checkbox root passes `field-complete` when at least one exported child column
is selected. Before assessment, `run_plan()` requires every detection column,
selected response column, checkbox child column, and branching dependency for
the frozen targets. A branching dependency may belong to an instrument with no
target of its own. Other fields exclusive to nontarget instruments are not
required.

## Compare assessments

Use `compare_reports(previous, current)` when the same project has been assessed
at two times or under different target plans. Both reports must retain
`details = TRUE` and versioned field-selection settings. Project identity,
structure fingerprints, and normalized field selection must match; target
membership and verification setup may change.

Two populations are shown by default:

- **Full scope** describes each report's complete assessment, including targets
  that entered or left the plan.
- **Shared targets** restricts each side to the exact record, event, instrument,
  repeat instrument, and repeat instance keys present in both plans. Stored
  outcomes are subset; branching logic is never reevaluated on a reduced export.

The following synthetic offline project has a Visit instrument at Baseline and
a repeating Diary at Follow up. Record `004` enters the Visit plan, while
record `001` leaves the Diary instance-2 plan. Previously unstarted targets
begin assessment, and reviewer `reviewer_a` verifies one remaining Diary field.

```{r comparison-project}
comparison_metadata <- data.frame(
  field_name = c("record_id", "visit_start", "visit_a", "visit_b",
                 "diary_start", "diary_a", "diary_b"),
  form_name = c(rep("visit", 4), rep("diary", 3)),
  field_type = "text",
  field_label = c("Record ID", "Visit started", "Visit A", "Visit B",
                  "Diary started", "Diary A", "Diary B"),
  required_field = c("", "", "y", "y", "", "y", "y")
)
comparison_rcon <- suppressWarnings(redcapAPI::offlineConnection(
  meta_data = comparison_metadata,
  project_info = data.frame(project_id = "77", is_longitudinal = "1",
                           has_repeating_instruments_or_events = "1"),
  arms = data.frame(arm_num = 1L, name = "Arm 1"),
  events = data.frame(event_id = c(101L, 102L), arm_num = 1L,
    unique_event_name = c("baseline_arm_1", "followup_arm_1"),
    event_name = c("Baseline", "Follow up")),
  instruments = data.frame(instrument_name = c("visit", "diary"),
                           instrument_label = c("Visit", "Diary")),
  mapping = data.frame(arm_num = 1L,
    unique_event_name = c("baseline_arm_1", "followup_arm_1"),
    form = c("visit", "diary")),
  repeat_instrument = data.frame(event_name = "followup_arm_1", form_name = "diary")
))
example_row <- function(id, event, instance = NA_integer_,
                        start = "yes", a = "entered", b = "entered") {
  diary <- event == "followup_arm_1"
  data.frame(record_id = id, redcap_event_name = event,
    redcap_repeat_instrument = if (diary) "diary" else NA_character_,
    redcap_repeat_instance = instance,
    visit_start = if (!diary) start else "",
    visit_a = if (!diary) a else "", visit_b = if (!diary) b else "",
    diary_start = if (diary) start else "",
    diary_a = if (diary) a else "", diary_b = if (diary) b else "")
}
previous_data <- dplyr::bind_rows(
  example_row("001", "baseline_arm_1", a = ""),
  example_row("002", "baseline_arm_1"),
  example_row("003", "baseline_arm_1", start = "", a = "", b = ""),
  example_row("001", "followup_arm_1", 2L, a = ""),
  example_row("002", "followup_arm_1", 2L, a = ""),
  example_row("003", "followup_arm_1", 1L)
)
current_data <- dplyr::bind_rows(
  example_row("001", "baseline_arm_1"),
  example_row("002", "baseline_arm_1"),
  example_row("003", "baseline_arm_1", a = ""),
  example_row("004", "baseline_arm_1", start = "", a = "", b = ""),
  example_row("002", "followup_arm_1", 2L, a = ""),
  example_row("003", "followup_arm_1", 2L, a = "")
)
example_schedule <- function(visit_ids, diary_ids) {
  dplyr::bind_rows(
    data.frame(record_id = visit_ids, instrument = "visit",
               redcap_event_name = "baseline_arm_1", repeat_instance = NA_integer_),
    data.frame(record_id = diary_ids, instrument = "diary",
               redcap_event_name = "followup_arm_1", repeat_instance = 2L)
  )
}
previous_plan <- plan_explicit(previous_data, comparison_rcon,
  example_schedule(c("001", "002", "003"), c("001", "002", "003")))
current_plan <- plan_explicit(current_data, comparison_rcon,
  example_schedule(c("001", "002", "003", "004"), c("002", "003")))
comparison_verified <- data.frame(
  project_id = "77", record = "002", event_id = "102", field_name = "diary_a",
  repeat_instrument = "diary", instance = 2L, ts = "2026-01-02 10:00:00",
  current_query_status = "VERIFIED", username = "reviewer_a"
)
previous <- run_plan(previous_plan, previous_data, comparison_rcon,
                     details = TRUE, progress = FALSE)
current <- run_plan(current_plan, current_data, comparison_rcon,
  verified = comparison_verified, verified_user = "reviewer_a",
  details = TRUE, progress = FALSE)
comparison <- compare_reports(previous, current)
```

### Validation summaries

Each row identifies population, raw event, instrument, repeat instrument,
exact repeat instance, registry level, and check. All four checks remain in
registry order, including not-applicable checks. Each side retains status,
reason, assessed/passed/failed counts, and pass/fail rates. Absent contexts have
false presence flags, zero counts, and missing status and rates.

```{r comparison-summary}
summary_rows <- get_summary(comparison, validation_check = "field-complete")
knitr::kable(summary_rows[, c(
  "population", "redcap_event_name", "instrument", "repeat_instance",
  "previous_failed", "previous_assessed", "current_failed", "current_assessed"
)])
```

For shared Baseline/Visit targets, field failures are **1/4 → 1/6**: the
denominator increases because record `003` now starts. The same target
population does not guarantee the same assessed field population. Field counts
represent field occurrences; a checkbox root counts once. The other three
checks count target outcomes within their stratum. Failures across these
different checks must not be added into one headline count.

### Event and instrument view

```{r comparison-instruments}
if (requireNamespace("flextable", quietly = TRUE) &&
    requireNamespace("glue", quietly = TRUE)) {
  flex_event_instruments(comparison)
}
```

Baseline/Visit incomplete instruments change **2/3 → 2/4** in Full scope,
versus **2/3 → 1/3** on Shared targets. Follow-up/Diary instance 2 has **3/3 →
1/2** incomplete instruments in Full scope and **2/2 → 1/2** on Shared targets.
Incomplete instruments count target outcomes, not field failures. Both sides
use the same threshold. Rate differences are percentage points.

Event headers deduplicate record/event contexts across instruments, repeating
rows show repeat-gate counts, and the All row sums instrument opportunities.
An absent context displays "Out of scope"; disjoint plans display "No shared
targets" in the shared section.

### Record and field changes

```{r comparison-changes}
knitr::kable(get_changes(comparison, instruments = "diary")[, c(
  "record_id", "repeat_instance", "validation_check", "field_name", "change", "reason"
)])
```

`get_changes()` retains failures present in either report. Exact target keys
plus check and field name identify each failure; leading-zero IDs are retained.
Added/removed targets take scope precedence. Within shared targets,
`newly_detected`, `still_missing`, `completed`, `verified`, and
`no_longer_assessed` distinguish new failure, persistent failure, direct pass,
verification, and branching/gate exclusion. Newly reached checks and loss of
verification have explicit reasons. Verification setup changes are flagged in
`comparison$verification` without claiming they caused an outcome change.

Each stratum reconciles previous failures to still-missing, completed,
verified, no-longer-assessed, and removed-scope failures; current failures
reconcile to still-missing, newly detected, and added-scope failures. Full-scope
summary sides agree with the source reports. Shared summaries contain only
retained-target transitions.

```{r comparison-filters, eval = FALSE}
get_summary(comparison, population = "shared", instruments = "visit")
get_changes(comparison, change = c("newly_detected", "still_missing"))
comparison$scope_changes  # Includes added/removed targets with no failures.
flexify(get_changes(comparison))
```

The comparison preserves plans, paired target results, normalized settings,
summary strata, transitions, scope changes, and verification mode/reviewer.
It excludes response values and raw verification history. Available current
links are preferred, with previous links as fallback. Existing saved reports
remain usable with report accessors and formatters; regenerate legacy or
nondetailed reports before comparing them.

## Verification

Verification is optional and intended for projects that use REDCap's Data
Resolution Workflow. Other projects can leave `verified` and
`verified_user` at their defaults.

### Enable the module and export history

The [Vanderbilt Data Quality API module](https://github.com/vanderbilt-redcap/data_quality_api)
must be enabled in the REDCap Control Center and in your project. Use an
authorized live `redcapApiConnection`. The default module prefix is
`"data_quality_api"`; check the module's configuration/link or ask your
REDCap administrator if your installation uses another prefix.

```{r verified-export, eval = FALSE}
verified <- export_data_quality(rcon)
# To restrict the request to selected records:
# verified <- export_data_quality(rcon, records = c("1", "2"))

verified_report <- run_plan(
  plan = explicit_plan,
  data = records,
  rcon = rcon,
  verified = verified,
  verified_user = "reviewer",
  details = TRUE
)
```

`records = NULL` retrieves the complete project history and can take time on
large projects. Export once and reuse the result across assessments.
`run_plan()` never retrieves history. There are no status or user retrieval
filters: later entries must remain available to determine whether an earlier
verification still applies.

The export is a plain data frame with one row per resolution, carrying its
parent issue's fields. Issues without resolutions contribute a row with
missing resolution values. All exported columns are character; timestamps
retain their source text, and JSON nulls become `NA_character_`. Additional
scalar module fields are retained. See `?export_data_quality` for the full
return schema, including empty results.

### Match history to the plan

Pass the returned table directly. No manual flattening, context normalization,
or filtering to the plan is needed. `run_plan()` uses these nine columns and
ignores extras:

| Column | Meaning |
|---|---|
| `project_id` | Project identity; every supplied row must match the plan's project. |
| `record` | Raw record ID; character leading zeros are preserved. |
| `event_id` | Internal event ID; mapped to a raw event name for longitudinal projects. A consistent positive classic event ID, or a missing value, maps to the eventless context. |
| `field_name` | Raw metadata field name. |
| `repeat_instrument` | Repeating instrument name, or missing when inapplicable. |
| `instance` | Actual repeating instance; native placeholder `1` becomes missing for a nonrepeating target. |
| `ts` | Resolution timestamp. Required for in-plan entries with a reviewer; text without an offset is interpreted as UTC. |
| `current_query_status` | Resolution status. Only the selected reviewer's latest exact `"VERIFIED"` can apply. |
| `username` | Resolution reviewer, matched exactly and case sensitively to `verified_user`. |

A complete zero-row table is valid. Existing normalized tables remain
accepted; `?run_plan` documents their storage and timestamp formats.

The table structure and project identity are checked first. Well-formed
contexts outside the plan are silently ignored before resolution values are
validated, including unknown records, events, instruments, and fields.
Malformed identifiers still raise errors. Evidence never expands the plan.
Issue-only rows and entries without an identifiable reviewer are quietly
ignored. Other in-plan entries require valid timestamps.

For the selected reviewer, rows are grouped by normalized record, event,
repeat instrument, repeat instance, and field. The latest timestamp wins.
Identical latest duplicates collapse; conflicting latest ties raise
`redcapmissing_error_verification`. If the latest status is missing, blank,
`"OPEN"`, `"CLOSED"`, or `"DEVERIFIED"`, no override applies. An older
`"VERIFIED"` is never substituted. Issue-level `query_status` and
`assigned_username` do not control assessment.

### Runnable offline example

This synthetic example uses three started records with a missing required
field. The history table represents selected columns of a flattened export:
record `1` has a verification, record `2` has a later reopening by the same
reviewer, and record `3` has an issue without any resolutions. A live workflow
obtains this table with `export_data_quality()`.

```{r verified-offline}
metadata <- data.frame(
  field_name = c("record_id", "started", "value"),
  form_name = "baseline",
  field_type = "text",
  field_label = c("Record ID", "Started", "Value"),
  required_field = c("y", "", "y")
)
project_info <- data.frame(
  project_id = "1",
  is_longitudinal = "0",
  has_repeating_instruments_or_events = "0"
)
rcon <- suppressWarnings(redcapAPI::offlineConnection(
  meta_data = metadata,
  project_info = project_info,
  repeat_instrument = redcapAPI::REDCAP_REPEAT_INSTRUMENT_STRUCTURE
))
records <- data.frame(
  record_id = c("1", "2", "3"),
  started = "yes",
  value = ""
)
verified <- data.frame(
  project_id = "1",
  record = c("1", "2", "2", "3"),
  event_id = "101",
  field_name = "value",
  repeat_instrument = NA_character_,
  instance = "1",
  status_id = c("10", "11", "11", "12"),
  res_id = c("20", "21", "22", NA_character_),
  ts = c(
    "2026-01-02 10:00:00", "2026-01-02 10:00:00",
    "2026-01-03 10:00:00", NA_character_
  ),
  current_query_status = c("VERIFIED", "VERIFIED", "OPEN", NA_character_),
  username = c("reviewer", "reviewer", "reviewer", NA_character_)
)

plan <- plan_from_data(records, rcon, "baseline")
before <- run_plan(plan, records, rcon, progress = FALSE)
after <- run_plan(
  plan, records, rcon,
  verified = verified, verified_user = "reviewer",
  details = TRUE, progress = FALSE
)
knitr::kable(data.frame(
  record_id = before$target_results$record_id,
  before = before$target_results$field_complete,
  after = after$target_results$field_complete
))
```

Record `1` changes from a failure to a pass. Records `2` and `3` remain
unresolved. Verification affects only a failed, selected, applicable field
whose target passes the upstream row and instrument-start checks.

```{r verified-details}
field_rows <- after$details[
  after$details$validation_check == "field-complete" &
    after$details$field_name %in% "value",
]
knitr::kable(field_rows[, c(
  "record_id", "raw_disposition", "verification_applied", "effective_disposition"
)])
knitr::kable(data.frame(
  audit_field = names(after$verification),
  value = unlist(after$verification, use.names = FALSE)
))
```

### Interpret the audit and resolve common problems

| Audit field | Meaning |
|---|---|
| `enabled` | `TRUE` when both verification arguments were supplied, including an empty table. |
| `verified_user` | The exact supplied username; `NA_character_` when disabled. |
| `input_rows` | All supplied rows, including issue-only and out-of-plan rows. |
| `user_rows` | All supplied rows whose username exactly matches the selected reviewer, including out-of-plan rows. |
| `latest_user_rows` | In-plan field contexts retained after latest-entry selection and duplicate collapse, including missing latest statuses. |
| `verified_rows` | Retained entries with exact `"VERIFIED"`. |
| `overrides_applied` | Failed field results actually changed to passes. |

The example has four input rows, three reviewer rows, two latest entries,
one latest verification, and one override. The report excludes raw workflow
history.

- **Export fails or returns malformed JSON:** Check the connection's export
  permissions, module enablement, and prefix. Structurally inconsistent
  history raises `redcapmissing_error_export` without including response
  contents in the error.
- **No reviewer rows:** Check the exact spelling and case of
  `verified_user`; it identifies the resolution's author, not its assignee.
- **Latest entry has no status:** No override applies. Review the complete
  history; do not remove that entry to reveal an older verification.
- **Verified rows but no override:** Check that the field is assessed, still
  fails, and passes the row, instrument-start, branching, and selection rules.

## Response missingness

Ordinary fields used by `instrument-started` and `field-complete` use the
response rule in this table. Checkbox roots use the selected child rule in
[Instrument detection and field selection](#instrument-detection-and-field-selection).

| Response value | Treatment |
|---|---|
| Typed R `NA`, factor `NA`, logical `NA`, numeric `NA`, or `NaN` | Missing. |
| Missing `Date` or `POSIXct` | Missing. |
| Empty character text or text containing only whitespace | Missing. |
| `Inf` or `-Inf` | Present. |
| Literal `"NA"`, `"N/A"`, `"NULL"`, `"."`, or `"-999"` | Present unless an earlier REDCap missing code process converted it. |

Physical row presence depends only on the structural key.

## Missing values, columns, and rows

The same R representation has a different result at each input:

| Representation | Structural identifier or schedule cell | Response field | Scalar or vector argument |
|---|---|---|---|
| Typed `NA` | Accepted for an inapplicable dimension; an applicable dimension requires a value. | Missing. | Accepted only where the argument documentation allows it; logical scalar arguments reject `NA`. |
| `NaN` | Error, including for an all-missing nullable structural column. | Missing. | Error for numeric structural values. |
| `Inf` or `-Inf` | Error for numeric identifiers and instances. | Present. | Error where finite values are required. |
| `""` or whitespace only | Converted to typed `NA` for an inapplicable dimension; an applicable dimension requires a value. | Missing. | Names and usernames require nonblank text. |
| `"NA"`, `"N/A"`, `"NULL"`, `"."`, or `"-999"` | Literal text subject to the identifier or schedule rules. | Present. | Literal text subject to the argument rules. |
| `NULL` | Unavailable as a table cell. | Unavailable as a table cell. | Selects omission only where the argument permits `NULL`. |
| `character(0)` | Unavailable as a table cell. | Unavailable as a table cell. | Represents no removal for `ignore_fields` and `exclude_types`; `instruments` requires at least one value. |
| Data frame with zero rows | Planner schedules require their exact schema and documented storage. A `verified` table requires each of its nine column names once; storage is checked when rows exist. | Supplies no responses. | `plan_from_data()` rejects zero row `data`; `plan_explicit()` and `run_plan()` accept correctly structured zero row `data`. |
| Absent column | An error when the project structure requires the column. | An error when a selected check or branching expression requires the column. | No argument value exists. |
| Absent row | Planning and running apply the input specific rules below. | A stored target remains; its `redcap_event_name`, `repeat_instrument`, and `repeat_instance` values determine the check statuses. | No argument value exists. |

Absent row rules are:

| Input | Result |
|---|---|
| Planner `data` | The crossing is unobserved. A schedule may add a target. |
| `extended_schedule` | The observed planning rule applies to that crossing. |
| `explicit_schedule` | The exact target is excluded. |
| Runtime `data` | A stored target remains. An absent longitudinal event fails `event-row-started`; an absent repeat instance after a present event fails `repeat-instance-row-started`; an absent classic target with `repeat_instance = NA_integer_` has both row checks marked `"not applicable"` and fails `instrument-started`. |
| `verified` | That field context receives no verification evidence. |

## Result components

`run_plan()` returns a `redcapmissing` object with these components:

| Component | Contents |
|---|---|
| `plan` | The validated `redcapmissing_plan` used for assessment. |
| `target_results` | One row per `assessible_targets` row, with source, four check statuses, field counts, and `field_applicability_reason`. |
| `summary` | Counts and rates grouped by event, instrument, repeat context, validation level, and validation check. |
| `missing` | Effective failures that remain unresolved after verification. |
| `verification` | Counts named `enabled`, `verified_user`, `input_rows`, `user_rows`, `latest_user_rows`, `verified_rows`, and `overrides_applied`. |
| `diagnostics` | The twelve stages with completion and elapsed time. |
| `details` | Detailed check and field rows when `details = TRUE`; `NULL` when `details = FALSE`. |
| `settings` | Versioned `required_fields`, `ignore_fields`, and `exclude_types`; empty selections are character vectors and selection order is normalized. |

`summary$status` is `"assessed"` or `"not applicable"`.
`summary$reason` is typed missing for assessed rows and contains the applicable
reason for a row with status `"not applicable"`. Rates are `NA_real_` when a
group has no assessed checks.

`missing` uses typed missing values for inapplicable event and repeat dimensions.
Its `url` column may contain a REDCap data entry URL when `rcon` supplies the
required URL, version, and event data.

### Detailed rows and stored values

With `details = TRUE`:

- `event-row-started`, `repeat-instance-row-started`, and
  `instrument-started` contribute one row per target.
- `field-complete` contributes one row for each field assessed for the target.
- A target with `field-complete` equal to `"not applicable"` or
  `"not reached"` contributes one target row for that check.
- Fields whose branching logic is false contribute no field row.

`raw_disposition` and `effective_disposition` use `"passed"`, `"failed"`,
`"not applicable"`, and `"not reached"`. Verification can change a raw
`"failed"` field row to an effective `"passed"` row.
`verification_applied` records that change.

For an assessed field, `branch_satisfied` is `TRUE`. Target check rows store
`NA` in this column. `reason` uses these exact values where they apply:

| Validation check | `reason` |
|---|---|
| `event-row-started` | `"not applicable for classic project"` |
| `repeat-instance-row-started` | `"not a repeating target"` |
| `field-complete` after field selection | `"no fields remain after field policy"` |
| `field-complete` after branching logic | `"no fields apply after branching logic"` |

Other detail rows store `NA_character_` in `reason`.

`value_summary` stores:

| Detail row | Stored value |
|---|---|
| Assessed ordinary field | The assessed value converted to character text. |
| Assessed checkbox root with selected choices | The selected exported checkbox child column names, separated by commas. |
| Assessed checkbox root with no selected choice | An empty character value. |
| Target check row | `NA_character_`. |

The report contains record IDs. It can also contain REDCap data entry URLs and,
when `details = TRUE`, assessed response values in `details$value_summary`.
Store the report under the same access and retention rules used for the REDCap
export.

With `details = FALSE`, `details` is `NULL`. `target_results`, `summary`,
`missing`, and `verification` have the same assessment result in both settings.

## Inspect and present results

```{r inspect, eval = FALSE}
get_summary(report)
get_missing(report)

flexify(get_summary(report))
flex_event_instruments(report)
flex_html(flexify(get_missing(report)))
```

Values within one accessor filter vector are alternatives. Multiple non-`NULL`
filters are combined by intersection. Matching uses exact raw values and is
case sensitive; duplicate requested values normalize to one value, and unknown,
missing, blank, or padded values error. Filtering preserves the stored report
row order.

When no rows match, `get_summary()` retains the names, order, and storage types
documented for `report$summary`; `get_missing()` retains those documented for
`report$missing`. The `redcapmissing_labels` attribute continues to contain
labels for every event and instrument represented by `report$plan`, including
labels absent from the returned rows. See `?get_summary`, `?get_missing`,
`?flex_event_instruments`, and `?flex_html` for their specific arguments and
return classes.

## Errors and recovery

Validation errors inherit from `redcapmissing_error`:

| Class | Boundary to inspect |
|---|---|
| `redcapmissing_error_argument` | Missing arguments, invalid scalar values, or invalid selection vectors. |
| `redcapmissing_error_schema` | Required columns, storage, identifier or repeat instance values, or duplicate physical row keys. This includes schedule value conversion. |
| `redcapmissing_error_project` | Project surfaces from `rcon`, project mappings, repeat settings, or branching logic. |
| `redcapmissing_error_schedule` | Schedule column names and order, missing or unknown selections, duplicate converted rows, or disallowed crossings. |
| `redcapmissing_error_plan` | Plan class, components, schema version, target order, fingerprint, or changed project structure. |
| `redcapmissing_error_verification` | Verification schema, values, target contexts, timestamps, or tied latest rows. |

Use the class and message to identify the failing input:

1. For a schema error, check required columns and their storage. For runtime
   `data`, export every structural column plus each detection, response,
   checkbox, and branching dependency column required by the frozen targets.
   Nontarget instruments impose no exclusive response-column requirement.
2. For a schedule error, check the exact column names and order, duplicates,
   instrument and event names, and the event repeat setting in `rcon`.
3. For a plan error after a REDCap structure change, create a new plan from the
   current `rcon`.
4. For a project error involving a reference to another event, review the
   source rows. The reference requires one row with a missing
   `redcap_repeat_instance` or, when no such row exists, one row with a
   positive repeat instance.
5. For a verification error, validate every row and resolve different rows
   tied at the latest timestamp for one field context.
6. For `redcapmissing_warning_empty_arm_extension`, confirm that planner
   `data` contains records in the scheduled event arm. The warning adds no
   targets for that arm.
7. For `redcapmissing_warning_undesignated_extension`, review the REDCap event
   designations for the named longitudinal instruments. They contributed no
   builder rows; no eventless longitudinal crossing was created.

After correcting the source input, call the constructor or `run_plan()` again.
