---
title: "Getting started with redcapmissing"
vignette: >
  %\VignetteIndexEntry{Getting started with redcapmissing}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`redcapmissing` builds branching-aware missingness reports for REDCap record exports.

This package is built on top of `redcapAPI`. In real workflows, callers create
an `rcon` object with `redcapAPI::redcapConnection()`, export records with
`redcapAPI::exportRecordsTyped()`, and then pass both objects into
`redcap_missing_report()`.

The examples below are intentionally synthetic so the vignette remains runnable
without live REDCap credentials or access to a specific REDCap project.

## A minimal synthetic REDCap example

```{r}
library(redcapmissing)

meta_row <- function(
  field_name,
  form_name,
  field_type = "text",
  field_label = field_name,
  choices = "",
  branching = "",
  required = ""
) {
  tibble::tibble(
    field_name = field_name,
    form_name = form_name,
    field_type = field_type,
    field_label = field_label,
    select_choices_or_calculations = choices,
    text_validation_type_or_show_slider_number = "",
    branching_logic = branching,
    required_field = required
  )
}

fake_rcon <- function(metadata, mapping = NULL, repeat_instrument_event = NULL) {
  list(
    metadata = function() metadata,
    mapping = function() mapping,
    mappings = function() mapping,
    repeatInstrumentEvent = function() repeat_instrument_event
  )
}
```

## Real `redcapAPI` workflow

In routine use, the package expects a `redcapAPI` connection and a typed export:

```{r, eval = FALSE}
rcon <- redcapAPI::redcapConnection(
  url = Sys.getenv("REDCAP_API_URL"),
  token = Sys.getenv("REDCAP_API_TOKEN")
)

records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)

report <- redcap_missing_report(
  data = records,
  rcon = rcon,
  form = "baseline_form"
)
```

## Required and branching-aware field checks

```{r}
metadata <- dplyr::bind_rows(
  meta_row("record_id", "baseline_form", required = "y"),
  meta_row("branch_flag", "baseline_form", field_type = "yesno", required = "y"),
  meta_row("required_note", "baseline_form", required = "y"),
  meta_row("conditional_note", "baseline_form", branching = "[branch_flag] = '1'", required = "y")
)

records <- tibble::tibble(
  record_id = c("r1", "r2"),
  branch_flag = c("1", "0"),
  required_note = c("entered", "entered"),
  conditional_note = c("", "")
)

report <- redcap_missing_report(
  data = records,
  rcon = fake_rcon(metadata),
  form = "baseline_form"
)

report$missing[, c("record_id", "field_name", "missing_scope")]
```

## Repeat expectations

```{r}
repeat_metadata <- dplyr::bind_rows(
  meta_row("record_id", "screen_form", required = "y"),
  meta_row("screen_started", "screen_form", required = "y"),
  meta_row("repeat_started", "repeat_form", required = "y"),
  meta_row("repeat_value", "repeat_form", required = "y")
)

mapping <- tibble::tibble(
  arm_num = c(1, 1),
  unique_event_name = c("baseline_event", "baseline_event"),
  form = c("screen_form", "repeat_form")
)

repeat_map <- tibble::tibble(
  event_name = "baseline_event",
  form_name = "repeat_form",
  custom_form_label = ""
)

repeat_records <- tibble::tibble(
  record_id = c("r1", "r2", "r2"),
  redcap_event_name = c("baseline_event", "baseline_event", "baseline_event"),
  redcap_repeat_instrument = c("repeat_form", "repeat_form", "repeat_form"),
  redcap_repeat_instance = c("1", "1", "2"),
  repeat_started = c("yes", "yes", "yes"),
  repeat_value = c("10", "20", "30")
)

repeat_report <- redcap_missing_report(
  data = repeat_records,
  rcon = fake_rcon(repeat_metadata, mapping = mapping, repeat_instrument_event = repeat_map),
  form = "repeat_form",
  expected_repeats = 2L
)

repeat_report$repeat_missing[, c("record_id", "redcap_repeat_instance", "missing_scope")]
```

## Summary tables

If optional formatting packages are installed, you can format the validation summary:

```{r, eval = FALSE}
redcap_missing_summary(report)
```

## Citation and acknowledgement

`redcapmissing` depends heavily on `redcapAPI`. Please cite `redcapAPI` when
`redcapmissing` is part of a REDCap workflow:

> Nutter B, Garbett S, Obregon S, Obadia T, Lehr M, High B, Lane S, Beasley W,
> Gray W, Kennedy N, Hsi-Nien T, Horner J, Stephens J, Beck C, Johnson B,
> Chase P, Tobias P (2026). *redcapAPI: Accessing data from REDCap projects
> using the API*. R package version 2.12.0.
> <https://doi.org/10.5281/zenodo.10564837>
