---
title: "Creating and running an assessment plan"
vignette: >
  %\VignetteIndexEntry{Creating and running an assessment plan}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(redcapmissing)
```

`redcapmissing` separates two decisions that are easy to blur together:

1. Which record, event, instrument, and exact repeat-instance crossings are
   Assessible?
2. For those frozen targets, which physical rows, instruments, and fields are
   complete?

Construct a `redcapmissing_plan` with `plan_from_data()` or `plan_explicit()`,
then evaluate it with `run_plan()`.

## Export records

In a live workflow, create a `redcapAPI::redcapConnection()` and export typed
records. Keep fields used in branching logic in REDCap code space and keep
checkbox and REDCap system columns raw.

```{r export, eval = FALSE}
records <- redcapAPI::exportRecordsTyped(
  rcon,
  cast = list(
    radio = redcapAPI::castCode,
    dropdown = redcapAPI::castCode,
    yesno = redcapAPI::castCode,
    truefalse = redcapAPI::castCode,
    checkbox = redcapAPI::castRaw,
    system = redcapAPI::castRaw
  )
)
```

The project record-ID field is discovered from `rcon`; it need not be named
`record_id` in `records`. Constructors inspect physical rows only. Blank
responses do not make an exported row unobserved.

## Planning functions and arguments

| Function | Argument | Required | Accepted values | Behavior |
|---|---|---:|---|---|
| Both | `data` | Yes | A data frame or tibble containing REDCap rows and the structural columns required by the project. | Supplies observed physical rows; only `plan_from_data()` uses them to establish the extension population. |
| Both | `rcon` | Yes | A REDCap connection or supported connection-like object exposing all project structure surfaces. | Supplies project identity, metadata, instruments, events, arms, mappings, and repeat configuration. |
| Both | `instruments` | Yes | A nonempty character vector of unique, nonmissing, nonblank, unpadded raw instrument names. | Selects the instrument domain and determines target ordering. |
| `plan_from_data()` | `extended_schedule` | No | `NULL`, a correctly typed zero-row table, or the exact three-column schedule below. | Adds sparse event/instance expectations without removing observed targets. |
| `plan_explicit()` | `explicit_schedule` | Yes | The exact four-column table below; `NULL` is invalid and a correctly typed zero-row table is valid. | Completely declares the record/instrument/event/instance targets; omitted rows are not assessed. |

## Observed planning with sparse extensions

`plan_from_data()` uses this definition:

```text
Assessible targets =
  rcon-allowable crossings
  INTERSECT
  (
    crossings observed in data
    UNION
    extended_schedule crossings
  )
```

```{r observed-plan, eval = FALSE}
observed_plan <- plan_from_data(
  data = records,
  rcon = rcon,
  instruments = c("demographics", "labs")
)
```

`instruments` is a nonempty vector of unique, unpadded, raw REDCap instrument
names. `extended_schedule = NULL` and a correctly typed zero-row schedule both
mean observed-only planning.

The extension schema is exact; column order and types are part of the contract:

| Column | Normalized type | Meaning |
|---|---|---|
| `instrument` | Character | Selected raw instrument name. |
| `redcap_event_name` | Character | Exact raw event; typed `NA` only for a classic project. |
| `repeat_instance` | Integer | Exact positive instance for a repeating context; otherwise `NA_integer_`. |

```{r extension, eval = FALSE}
extended_schedule <- data.frame(
  instrument = c("labs", "labs"),
  redcap_event_name = c("month_3_arm_1", "month_3_arm_1"),
  repeat_instance = c(1L, 2L)
)

extended_plan <- plan_from_data(
  records,
  rcon,
  instruments = c("demographics", "labs"),
  extended_schedule = extended_schedule
)
```

Each extension row adds an exact instrument-event-instance expectation for all
records observed in the applicable arm. A classic extension expands across all
observed records. An extension into an arm with no observed records warns once
and adds no target. Omitted extension rows never remove observed targets.

## Authoritative explicit planning

`plan_explicit()` uses a complete record-level schedule:

```text
Assessible targets =
  rcon-allowable crossings
  INTERSECT
  explicit_schedule crossings
```

Its exact schema adds `record_id` at the beginning:

| Column | Normalized type | Meaning |
|---|---|---|
| `record_id` | Character | Exact record to assess; it may be absent from `data`. |
| `instrument` | Character | Selected raw instrument name. |
| `redcap_event_name` | Character | Exact raw event; typed `NA` only for a classic project. |
| `repeat_instance` | Integer | Exact positive instance for a repeating context; otherwise `NA_integer_`. |

```{r explicit-plan, eval = FALSE}
explicit_schedule <- data.frame(
  record_id = c("101", "101", "102"),
  instrument = c("demographics", "labs", "demographics"),
  redcap_event_name = c(
    "baseline_arm_1", "month_3_arm_1", "baseline_arm_1"
  ),
  repeat_instance = c(NA_integer_, 2L, NA_integer_)
)

explicit_plan <- plan_explicit(
  data = records,
  rcon = rcon,
  instruments = c("demographics", "labs"),
  explicit_schedule = explicit_schedule
)
```

No row expansion occurs. Each schedule row creates only that exact target.
Observed crossings absent from the schedule are excluded. A full-schema
zero-row schedule means assess nothing; `NULL` is invalid. A record absent from
`data` remains a target and can fail a physical-row check when the plan runs.

The key schedule distinction is therefore:

| Absence | `extended_schedule` | `explicit_schedule` |
|---|---|---|
| Instrument/event/instance row omitted | Make no extension; assess if allowable and observed. | Do not assess that crossing. |
| Crossing absent from planner data | An extension can add it for observed applicable records. | Its exact schedule row can add it, including for an absent record. |
| `NULL` | Valid observed-only planning. | Error. |
| Correct zero-row table | Valid observed-only planning. | Valid plan with zero targets. |

## Classic, longitudinal, and repeat contexts

In classic projects, `redcap_event_name` is structurally inapplicable and is
normalized to `NA_character_`. Schedule event cells may be typed missing or
blank; nonblank event names error.

Longitudinal data requires a nonblank raw `redcap_event_name` on every physical
row. Extensions expand only across records observed in that event's arm. An
explicit record observed in one arm cannot be scheduled into a contradictory
arm.

Repeating projects require `redcap_repeat_instrument` and
`redcap_repeat_instance` together in record data. The instrument column is:

- the raw instrument name for a repeating-instrument row;
- structurally `NA` for a regular row; and
- structurally `NA` for a repeating-event row.

The instance is a positive integer for every repeating row and structurally
`NA_integer_` for a regular row. An instrument can be regular at one event and
repeating at another:

```{r mixed-repeat, eval = FALSE}
mixed_schedule <- data.frame(
  instrument = c("labs", "labs", "labs"),
  redcap_event_name = c(
    "baseline_arm_1", "month_3_arm_1", "month_3_arm_1"
  ),
  repeat_instance = c(NA_integer_, 1L, 2L)
)
```

The first row is the regular baseline target. The next two rows are exact
month-3 repeat targets; instance `2L` does not imply instance 1.

## Plan contents

Both constructors return the same object with exactly these top-level
components:

```text
schema_version
construction
instruments
assessible_targets
project
structure_fingerprint
```

`construction` is `"from_data"` or `"explicit"`. The target table contains
canonical `record_id`, `instrument`, `redcap_event_name`, `repeat_instrument`,
`repeat_instance`, and `target_source`. Provenance is `"observed"`,
`"extended"`, `"observed+extended"`, or `"explicit"`.

Targets are unique and deterministically ordered. The structure fingerprint is
based on canonicalized project identity, metadata, instruments, arms, events,
mappings, and repeat configuration. The plan retains neither input records nor
a live connection.

## Running the plan

```{r run, eval = FALSE}
report <- run_plan(
  plan = extended_plan,
  data = records,
  rcon = rcon,
  required_fields = TRUE,
  ignore_fields = NULL,
  exclude_types = "descriptive",
  verified = NULL,
  verified_user = NULL,
  details = FALSE,
  progress = interactive()
)
```

| Argument | Required | Default | Accepted values | Purpose |
|---|---:|---|---|---|
| `plan` | Yes | None | A validated `redcapmissing_plan` from `plan_from_data()` or `plan_explicit()`. | Supplies frozen Assessible targets, selected instruments, provenance, and the project fingerprint. |
| `data` | Yes | None | A data frame or tibble with structural rows and every response/dependency column required by the selected checks. | Supplies the current REDCap rows and values to evaluate. |
| `rcon` | Yes | None | The same unchanged REDCap connection or supported connection-like object used to describe the plan's project. | Supplies current project identity and structure. |
| `required_fields` | No | `TRUE` | One nonmissing logical value. | Starts field-complete selection with required-only fields (`TRUE`) or all instrument fields (`FALSE`). |
| `ignore_fields` | No | `NULL` | `NULL`, `character(0)`, or unique nonblank raw metadata field names. | Removes named fields from field-complete assessment. |
| `exclude_types` | No | `"descriptive"` | `NULL`, `character(0)`, or unique nonblank metadata field types. | Removes field types from field-complete assessment. |
| `verified` | No | `NULL` | `NULL`, a complete zero-row template, or a data frame/tibble with the nine required columns below. | Supplies field-level verification evidence and must be paired with `verified_user`. |
| `verified_user` | No | `NULL` | `NULL` or one nonblank, unpadded character username. | Selects the exact case-sensitive reviewer and must be paired with `verified`. |
| `details` | No | `FALSE` | One nonmissing logical value. | Retains or omits detailed validation rows without changing assessment results. |
| `progress` | No | `interactive()` | One nonmissing logical value. | Enables or disables twelve-stage progress reporting. |

`run_plan()` accepts a newer record snapshot from the same structurally
unchanged project. Targets remain frozen: added unplanned rows cannot create
new targets, and an absent planned row fails its applicable gate.

The runner records these stages in order:

| Stage | Operation |
|---:|---|
| 1 | Validate plan, data, and `rcon`. |
| 2 | Validate and normalize verification. |
| 3 | Resolve instrument-start fields. |
| 4 | Resolve field-complete fields. |
| 5 | Join Assessible targets to physical rows. |
| 6 | Run `event-row-started`. |
| 7 | Run `repeat-instance-row-started`. |
| 8 | Run `instrument-started`. |
| 9 | Run raw `field-complete`. |
| 10 | Apply verification. |
| 11 | Aggregate effective results. |
| 12 | Construct the report. |

Every longitudinal target first receives an event gate. Every repeating target
receives a repeat-instance gate after the event gate. Classic event gates and
regular repeat gates are not applicable. A failed event or repeat gate makes
downstream checks not reached. Field completeness runs only after the
instrument-started check passes.

```{r registry}
registry()
```

Report validation levels are `event:instrument` and
`event:instrument:instance`.

## Instrument-started and field policy

Instrument-started asks whether data entry began. Its stable detection set is
all data-entry metadata fields on the selected instrument except the project
record ID and `descriptive` or `calc` fields. Checkbox roots expand to exported
child columns. A selected instrument with no usable detection fields errors.

Field-complete asks whether fields selected by policy are complete. The exact
policy order is:

```text
all fields on the selected instrument
-> required-only filter when required_fields = TRUE
-> exclude_types removal
-> ignore_fields removal
```

`ignore_fields` uses raw metadata field names, including checkbox roots rather
than exported checkbox-child names. Unknown, duplicate, blank, padded, or
irrelevant field names and types error.

These three arguments affect only field-complete checks. An optional, ignored,
or excluded-type field may still establish that the instrument started. If
policy leaves no fields, the field-complete status is `not applicable`, its
reason is `no assessible fields after field policy`, and rates are `NA_real_`.

Branching logic determines whether each selected field applies in its exact
context. Checkbox roots are complete when at least one exported child choice is
selected. Missing detection, selected response, checkbox-child, or branching
dependency columns error before assessment; an unexported field is never
silently treated as blank.

## Verification

`verified` and `verified_user` must be both supplied or both `NULL`. A supplied
verification table requires these columns; extra columns are ignored:

| Column | Accepted storage and values | Accepted missing values | Internal normalization |
|---|---|---|---|
| `project_id` | Nonblank character, integer, or finite whole-number double matching the project. | None. | Canonical character project ID. |
| `record` | Nonblank character, factor, integer, or finite double. | None. | Character record ID; factor labels are preserved. |
| `event_id` | Canonical positive integer as character, integer, or whole-number double; classic projects also accept an entirely missing logical column. | Typed `NA`, blank, or whitespace only where the classic event dimension is inapplicable. | Character event ID and mapped raw event name, or `NA_character_` where inapplicable. |
| `field_name` | Nonblank, unpadded character raw field name. | None. | Character. |
| `repeat_instrument` | Character raw instrument; an entirely missing logical, integer, or double column is accepted where inapplicable. | Typed `NA`, blank, or whitespace only for regular and repeating-event contexts. | Character instrument or `NA_character_`. |
| `instance` | Canonical positive integer as character, integer, or whole-number double; an entirely missing logical column is accepted where inapplicable. | Typed `NA`, blank, or whitespace only for regular contexts. | Positive integer or `NA_integer_`. |
| `ts` | Nonmissing `POSIXct` or documented REDCap/ISO-8601 character timestamp. | None. | `POSIXct` in UTC; timezone-less text is UTC. |
| `current_query_status` | Nonblank, unpadded character text. | None. | Case-preserved character; only exact `"VERIFIED"` can apply. |
| `username` | Nonblank, unpadded character text. | None. | Case-preserved character matched exactly to `verified_user`. |

```{r verified, eval = FALSE}
verified_report <- run_plan(
  explicit_plan,
  records,
  rcon,
  verified = quality_rows,
  verified_user = "reviewer_username",
  details = TRUE
)
```

Every row is validated before filtering. For the exact case-sensitive user, the
latest timestamp per normalized field context wins. Identical latest ties
collapse; conflicting latest ties error. Only exact `"VERIFIED"` changes an
otherwise-failing, already-assessed field-complete result. It cannot create a
target, bypass a gate, start an instrument, affect an already-passing field, or
restore a field removed by policy or branching.

The result stores verification audit counts, never the supplied rows. With
`details = TRUE`, detailed field rows include raw disposition, whether
verification applied, and effective disposition. Compact and detailed modes
produce the same targets, summary, and missing results.

## Missing, blank, and absent values

The same representation has different meanings at different exposure points:

| Representation | Structural identifier or schedule | Response field | Scalar/vector argument |
|---|---|---|---|
| Typed `NA` | Valid only where that dimension is inapplicable; otherwise error. | Missing. | Valid only when explicitly documented; logical scalar `NA` errors. |
| `NaN` | Error. | Missing. | Error where numeric values are accepted structurally. |
| `Inf`, `-Inf` | Error for numeric identifiers and instances. | Not automatically missing. | Error where finite values are required. |
| `""` or whitespace-only | Structural `NA` only where inapplicable; otherwise error. | Missing. | Blank names and users error. |
| `"NA"`, `"N/A"`, `"NULL"`, `"."`, `"-999"` | Literal text, validated as such. | Ordinary nonmissing text. | Literal text, validated as such. |
| `NULL` | Not a cell value. | Not a cell value. | Omitted/disabled only where documented. |
| `character(0)` | Not a dataset cell. | Not a dataset cell. | No exclusion for field vectors; invalid for required instrument selection. |
| Zero-row data frame | Valid only with the complete required schema. | No responses. | Observed planner data cannot be zero-row; explicit planner/runtime data may be. |
| Absent column | Error when required by project structure or evaluation. | Error when required for a check. | Not applicable. |
| Absent row | Meaning depends on the input table. | A planned physical row can fail. | Not applicable. |

Record IDs accept character, factor, integer, or finite double and normalize to
character. Character leading zeros are preserved. Missing, blank, padded,
`NaN`, and infinite IDs error.

Repeat instances accept positive integer, whole-number double, or canonical
character digits and normalize to integer. Leading-zero strings, zero,
negative, decimal, missing repeating instances, noncanonical text, `NaN`,
infinity, and overflow error.

Physical row identity is unique on record ID, event, repeating instrument, and
instance. Normalization collisions produce duplicate-key errors; rows are never
silently discarded or identifiers trimmed.

For absent rows specifically:

- planner `data`: a crossing is unobserved but may be scheduled;
- `extended_schedule`: make no extension and retain any allowable observed
  crossing;
- `explicit_schedule`: do not assess that exact crossing;
- runtime `data`: the frozen target remains and may fail a physical-row gate;
- `verified`: no evidence exists for that exact field context.

## Results and presentation

A `redcapmissing` object has exactly:

```text
plan
target_results
summary
missing
verification
diagnostics
details
```

`target_results` has one row per Assessible target and records provenance, all
four check statuses, field counts, and any field-applicability reason. Statuses
are `passed`, `failed`, `not applicable`, and `not reached`.

`summary` aggregates assessed, passed, and failed counts and exposes
applicability `status` and `reason`. `missing` contains effective unresolved
failures only. Structural columns use typed missing values, not blank-string
placeholders.

```{r inspect, eval = FALSE}
get_summary(report)
get_summary(report, instruments = "labs")
get_missing(report, validation_check = "field-complete")

flexify(get_summary(report))
flex_event_instruments(report, missing_threshold = 0.1)
flex_html(flexify(get_missing(report)))
```

Accessor filters are exact and case-sensitive. Their optional label attribute
contains named `events` and `instruments` vectors for presentation.